%% refman.latex %%

%% Copyright (C) 2000, 2001, 2002, 2003, 2004, 2005 Andrew Makhorin,
%% Department for Applied Informatics, Moscow Aviation Institute,
%% Moscow, Russia. All rights reserved. E-mail: <mao@mai2.rcnet.ru>.
%%
%% This file is part of GLPK (GNU Linear Programming Kit).
%%
%% GLPK is free software; you can redistribute it and/or modify it
%% under the terms of the GNU General Public License as published by
%% the Free Software Foundation; either version 2, or (at your option)
%% any later version.
%%
%% GLPK is distributed in the hope that it will be useful, but WITHOUT
%% ANY WARRANTY; without even the implied warranty of MERCHANTABILITY
%% or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public
%% License for more details.
%%
%% You should have received a copy of the GNU General Public License
%% along with GLPK; see the file COPYING. If not, write to the Free
%% Software Foundation, Inc., 59 Temple Place - Suite 330, Boston, MA
%% 02111-1307, USA.

\documentclass[a4paper,11pt,twoside,draft]{report}

\setlength{\paperwidth}{210mm}
\setlength{\paperheight}{297mm}
\setlength{\hoffset}{-1in}
\addtolength{\hoffset}{5mm}
\setlength{\oddsidemargin}{25mm}
\setlength{\voffset}{-1in}
\addtolength{\voffset}{20mm}
\setlength{\topmargin}{0mm}
\setlength{\headheight}{3mm}
\setlength{\headsep}{7mm}
\setlength{\textwidth}{150mm}
\setlength{\textheight}{230mm}
\setlength{\parindent}{18pt}

\pagestyle{myheadings}
\sloppy

\begin{document}

\begin{titlepage}
\begin{center}

\vspace*{2.5in}

\begin{huge}
{\bf GNU Linear Programming Kit}
\end{huge}

\vspace{0.5in}

\begin{LARGE}
{\bf Reference Manual}
\end{LARGE}

\vspace{0.5in}

\begin{LARGE}
{\bf Version 4.8}
\end{LARGE}

\vspace{0.5in}
\begin{Large}
(Draft Edition, January 2005)
\end{Large}
\end{center}

\newpage
\vspace*{1in}
\vfill

\noindent
The GLPK package is part of the GNU Project released under the aegis of
GNU.

\medskip \noindent
Copyright \copyright{} 2000, 2001, 2002, 2003, 2004, 2005 Andrew
Makhorin, Department for Applied Informatics, Moscow Aviation Institute,
Moscow, Russia. All rights reserved.

\medskip \noindent
Free Software Foundation, Inc., 59 Temple Place --- Suite 330, Boston,
MA 02111, USA.

\medskip \noindent
Permission is granted to make and distribute verbatim copies of this
manual provided the copyright notice and this permission notice are
preserved on all copies.

\medskip \noindent
Permission is granted to copy and distribute modified versions of this
manual under the conditions for verbatim copying, provided also that the
entire resulting derived work is distributed under the terms of
a permission notice identical to this one.

\medskip \noindent
Permission is granted to copy and distribute translations of this manual
into another language, under the above conditions for modified versions.
\end{titlepage}

\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}

GLPK (\underline{G}NU \underline{L}inear \underline{P}rogramming
\underline{K}it) is a set of routines written in the ANSI C programming
language and organized in the form of a callable library. It is intended
for solving linear programming (LP), mixed integer programming (MIP),
and other related problems.

\section{LP Problem}
\label{seclp}

GLPK assumes the following formulation of {\it linear programming (LP)}
problem:

\medskip

\noindent
\hspace{1in} minimize (or maximize)
$$Z = c_1x_{m+1} + c_2x_{m+2} + \dots + c_nx_{m+n} + c_0 \eqno (1.1)$$
\hspace{1in} subject to linear constraints
$$
\begin{array}{c}
\nonumber x_1 = a_{11}x_{m+1} + a_{12}x_{m+2} + \dots + a_{1n}x_{m+n} \\
\nonumber x_2 = a_{21}x_{m+1} + a_{22}x_{m+2} + \dots + a_{2n}x_{m+n} \\
\nonumber \dots \dots \dots \dots \dots \dots \\
\nonumber x_m = a_{m1}x_{m+1} + a_{m2}x_{m+2} + \dots + a_{mn}x_{m+n} \\
\end{array} \eqno (1.2)
$$
\hspace{1in} and bounds of variables
$$
\begin{array}{c}
\nonumber l_1 \leq x_1 \leq u_1 \\
\nonumber l_2 \leq x_2 \leq u_2 \\
\nonumber \dots \dots \dots \\
\nonumber l_{m+n} \leq x_{m+n} \leq u_{m+n} \\
\end{array} \eqno (1.3)
$$
where: $x_1, x_2, \dots, x_m$ --- auxiliary variables;
$x_{m+1}, x_{m+2}, \dots, x_{m+n}$ --- structural variables;
$Z$ --- objective function;
$c_1, c_2, \dots, c_n$ --- objective coefficients;
$c_0$~---~constant term (``shift'') of the objective function;
$a_{11}, a_{12}, \dots, a_{mn}$ --- constraint coefficients;
$l_1, l_2, \dots, l_{m+n}$ --- lower bounds of variables;
$u_1, u_2, \dots, u_{m+n}$ --- upper bounds of variables.

Auxiliary variables are also called {\it rows}, because they correspond
to rows of the constraint matrix (i.e. a matrix built of the constraint
coefficients). Analogously, structural variables are also called
{\it columns}, because they correspond to columns of the constraint
matrix.

Bounds of variables can be finite as well as infinite. Besides, lower
and upper bounds can be equal to each other. Thus, the following types
of variables are possible:
\begin{center}
\begin{tabular}{r@{}c@{}ll}
\multicolumn{3}{c}{Bounds of variable} & Type of variable \\
\hline
$-\infty <$ &$\ x_k\ $& $< +\infty$ & Free (unbounded) variable \\
$l_k \leq$ &$\ x_k\ $& $< +\infty$  & Variable with lower bound \\
$-\infty <$ &$\ x_k\ $& $\leq u_k$  & Variable with upper bound \\
$l_k \leq$ &$\ x_k\ $& $\leq u_k$   & Double-bounded variable \\
$l_k =$ &$\ x_k\ $& $= u_k$         & Fixed variable \\
\end{tabular}
\end{center}
\noindent
Note that the types of variables shown above are applicable to
structural as well as to auxiliary variables.

To solve the LP problem (1.1)---(1.3) is to find such values of all
structural and auxiliary variables, which:

a) satisfy to all the linear constraints (1.2), and

b) are within their bounds (1.3), and

c) provide a smallest (in the case of minimization) or a largest (in the
case of maximization) value of the objective function (1.1).

For solving LP problems GLPK uses a well known numerical procedure
called {\it the simplex method}. The simplex method performs iterations,
where on each iteration it transforms the original system of equaility
constraints (1.2) resolving them through different sets of variables to
an equivalent system called {\it the simplex table} (or sometimes
{\it the simplex tableau}), which has the following form:
$$
\begin{array}{c}
\nonumber \;\;\;\;\;\:Z =
   \;\;d_1(x_N)_1 + \;\;d_2(x_N)_2 + \dots + \;\;d_n(x_N)_n \\
\nonumber (x_B)_1 = \alpha_{11}(x_N)_1 + \alpha_{12}(x_N)_2 + \dots +
   \alpha_{1n}(x_N)_n \\
\nonumber (x_B)_2 = \alpha_{21}(x_N)_1 + \alpha_{22}(x_N)_2 + \dots +
   \alpha_{2n}(x_N)_n \\
\nonumber \dots \dots \dots \dots \dots \dots \\
\nonumber (x_B)_m = \alpha_{m1}(x_N)_1 + \alpha_{m2}(x_N)_2 + \dots +
   \alpha_{mn}(x_N)_n \\
\end{array} \eqno (1.4)
$$
where: $(x_B)_1, (x_B)_2, \dots, (x_B)_m$ --- basic variables;
$(x_N)_1, (x_N)_2, \dots, (x_N)_n$ --- non-basic variables;
$d_1, d_2, \dots, d_n$ --- reduced costs;
$\alpha_{11}, \alpha_{12}, \dots, \alpha_{mn}$ --- coefficients of the
simplex table. (May note that the original LP problem (1.1)---(1.3) also
has the form of a simplex table, where all equalities are resolved
through auxiliary variables.)

From the linear programming theory it is well known that if an optimal
solution of the LP problem (1.1)---(1.3) exists, it can always be
written in the form (1.4), where non-basic variables are set on their
bounds while values of the objective function and basic variables are
determined by the corresponding equalities of the simplex table.

A set of values of all basic and non-basic variables determined by the
simplex table is called {\it basic solution}. If all basic variables are
within their bounds, the basic solution is called {\it (primal)
feasible}, otherwise it is called {\it (primal) infeasible}. A feasible
basic solution, which provides a smallest (in case of minimization) or
a largest (in case of maximization) value of the objective function is
called {\it optimal}. Therefore, for solving LP problem the simplex
method tries to find its optimal basic solution.

Primal feasibility of some basic solution may be stated by simple
checking if all basic variables are within their bounds. Basic solution
is optimal if additionally the following optimality conditions are
satisfied for all non-basic variables:
\begin{center}
\begin{tabular}{lcc}
Status of $(x_N)_j$ & Minimization & Maximization \\
\hline
$(x_N)_j$ is free & $d_j = 0$ & $d_j = 0$ \\
$(x_N)_j$ is on its lower bound & $d_j \geq 0$ & $d_j \leq 0$ \\
$(x_N)_j$ is on its upper bound & $d_j \leq 0$ & $d_j \geq 0$ \\
\end{tabular}
\end{center}
In other words, basic solution is optimal if there is no non-basic
variable, which changing in the feasible direction (i.e. increasing if
it is free or on its lower bound, or decreasing if it is free or on its
upper bound) can improve (i.e. decrease in case of minimization or
increase in case of maximization) the objective function.

If all non-basic variables satisfy to the optimality conditions shown
above (independently on whether basic variables are within their bounds
or not), the basic solution is called {\it dual feasible}, otherwise it
is called {\it dual infeasible}.

It may happen that some LP problem has no primal feasible solution due
to incorrect formulation --- this means that its constraints conflict
with each other. It also may happen that some LP problem has unbounded
solution again due to incorrect formulation --- this means that some
non-basic variable can improve the objective function, i.e. the
optimality conditions are violated, and at the same time this variable
can infinitely change in the feasible direction meeting no resistance
from basic variables. (May note that in the latter case the LP problem
has no dual feasible solution.)

\section{MIP Problem}

{\it Mixed integer linear programming (MIP)} problem is LP problem in
which some variables are additionally required to be integer.

GLPK assumes that MIP problem has the same formulation as ordinary
(pure) LP problem (1.1)---(1.3), i.e. includes auxiliary and structural
variables, which may have lower and/or upper bounds. However, in case of
MIP problem some variables may be required to be integer. This
additional constraint means that a value of each {\it integer variable}
must be only integer number. (Should note that GLPK allows only
structural variables to be of integer kind.)

\section{Brief Example}

In order to understand what GLPK is from the user's standpoint, consider
the following simple LP problem:

\noindent
\hspace{1in} maximize
$$Z = 10 x_1 + 6 x_2 + 4 x_3$$
\hspace{1in} subject to
$$
\begin{array}{r@{}r@{}r@{}l}
\nonumber    x_1 &\ + \;\:x_2 &\ + \;\:x_3 &\ \leq 100 \\
\nonumber 10 x_1 &\ + 4 x_2 &\ + 5 x_3 &\ \leq 600 \\
\nonumber  2 x_1 &\ + 2 x_2 &\ + 6 x_3 &\ \leq 300 \\
\end{array}
$$
\hspace{1in} where all variables are non-negative
$$x_1 \geq 0, \ x_2 \geq 0, \ x_3 \geq 0$$

At first this LP problem should be transformed to the standard form
(1.1)---(1.3). This can be easily done by introducing auxiliary
variables, by one for each original inequality constraint. Thus, the
problem can be reformulated as follows:

\pagebreak

\noindent
\hspace{1in} maximize
$$Z = 10 x_1 + 6 x_2 + 4 x_3$$
\hspace{1in} subject to
$$
\begin{array}{r@{}l@{}l@{}l}
\nonumber p = &\;\:\;\:x_1 &+\;\:x_2 &+\;\:x_3 \\
\nonumber q = &10 x_1 &+ 4 x_2 &+ 5 x_3 \\
\nonumber r = &\;\:2  x_1 &+ 2 x_2 &+ 6 x_3 \\
\end{array}
$$
\hspace{1in} and bounds of variables
$$
\begin{array}{ccc}
\nonumber -\infty < p \leq 100 && 0 \leq x_1 < +\infty \\
\nonumber -\infty < q \leq 600 && 0 \leq x_2 < +\infty \\
\nonumber -\infty < r \leq 300 && 0 \leq x_3 < +\infty \\
\end{array}
$$
where $p, q, r$ are auxiliary variables (rows), and $x_1, x_2, x_3$ are
structural variables (columns).

The example C program shown below uses GLPK API routines in order to
solve this LP problem.

\begin{verbatim}
/* sample.c */

#include <stdio.h>
#include <stdlib.h>
#include "glpk.h"

int main(void)
{     LPX *lp;
      int ia[1+1000], ja[1+1000];
      double ar[1+1000], Z, x1, x2, x3;
s1:   lp = lpx_create_prob();
s2:   lpx_set_prob_name(lp, "sample");
s3:   lpx_set_obj_dir(lp, LPX_MAX);
s4:   lpx_add_rows(lp, 3);
s5:   lpx_set_row_name(lp, 1, "p");
s6:   lpx_set_row_bnds(lp, 1, LPX_UP, 0.0, 100.0);
s7:   lpx_set_row_name(lp, 2, "q");
s8:   lpx_set_row_bnds(lp, 2, LPX_UP, 0.0, 600.0);
s9:   lpx_set_row_name(lp, 3, "r");
s10:  lpx_set_row_bnds(lp, 3, LPX_UP, 0.0, 300.0);
s11:  lpx_add_cols(lp, 3);
s12:  lpx_set_col_name(lp, 1, "x1");
s13:  lpx_set_col_bnds(lp, 1, LPX_LO, 0.0, 0.0);
s14:  lpx_set_obj_coef(lp, 1, 10.0);
s15:  lpx_set_col_name(lp, 2, "x2");
s16:  lpx_set_col_bnds(lp, 2, LPX_LO, 0.0, 0.0);
s17:  lpx_set_obj_coef(lp, 2, 6.0);
s18:  lpx_set_col_name(lp, 3, "x3");
s19:  lpx_set_col_bnds(lp, 3, LPX_LO, 0.0, 0.0);
s20:  lpx_set_obj_coef(lp, 3, 4.0);
s21:  ia[1] = 1, ja[1] = 1, ar[1] =  1.0; /* a[1,1] =  1 */
s22:  ia[2] = 1, ja[2] = 2, ar[2] =  1.0; /* a[1,2] =  1 */
s23:  ia[3] = 1, ja[3] = 3, ar[3] =  1.0; /* a[1,3] =  1 */
s24:  ia[4] = 2, ja[4] = 1, ar[4] = 10.0; /* a[2,1] = 10 */
s25:  ia[5] = 3, ja[5] = 1, ar[5] =  2.0; /* a[3,1] =  2 */
s26:  ia[6] = 2, ja[6] = 2, ar[6] =  4.0; /* a[2,2] =  4 */
s27:  ia[7] = 3, ja[7] = 2, ar[7] =  2.0; /* a[3,2] =  2 */
s28:  ia[8] = 2, ja[8] = 3, ar[8] =  5.0; /* a[2,3] =  5 */
s29:  ia[9] = 3, ja[9] = 3, ar[9] =  6.0; /* a[3,3] =  6 */
s30:  lpx_load_matrix(lp, 9, ia, ja, ar);
s31:  lpx_simplex(lp);
s32:  Z = lpx_get_obj_val(lp);
s33:  x1 = lpx_get_col_prim(lp, 1);
s34:  x2 = lpx_get_col_prim(lp, 2);
s35:  x3 = lpx_get_col_prim(lp, 3);
s36:  printf("\nZ = %g; x1 = %g; x2 = %g; x3 = %g\n", Z, x1, x2, x3);
s37:  lpx_delete_prob(lp);
      return 0;
}

/* eof */
\end{verbatim}

The statement \verb|s1| creates a problem object using the routine
\verb|lpx_create_prob|. Being created the object is initially empty.
The statement \verb|s2| assigns a symbolic name to the problem object.

The statement \verb|s3| calls the routine \verb|lpx_set_obj_dir| in
order to set the optimization direction flag, where \verb|LPX_MAX| means
maximization.

The statement \verb|s4| adds three rows to the problem object.

The statement \verb|s5| assigns the symbolic name `\verb|p|' to the
first row, and the statement \verb|s6| sets the type and bounds of the
first row, where \verb|LPX_UP| means that the row has an upper bound.
The statements \verb|s7|, \verb|s8|, \verb|s9|, \verb|s10| are used in
the same way in order to assign the symbolic names `\verb|q|' and
`\verb|r|' to the second and third rows and set their types and bounds.

The statement \verb|s11| adds three columns to the problem object.

The statement \verb|s12| assigns the symbolic name `\verb|x1|' to the
first column, the statement \verb|s13| sets the type and bounds of the
first column, where \verb|LPX_LO| means that the column has an lower
bound, and the statement \verb|s14| sets the objective coefficient for
the first column. The statements \verb|s15|---\verb|s20| are used in the
same way in order to assign the symbolic names `\verb|x2|' and
`\verb|x3|' to the second and third columns and set their types, bounds,
and objective coefficients.

The statements \verb|s21|---\verb|s29| prepare non-zero elements of the
constraint matrix (i.e. constraint coefficients). Row indices of each
element are stored in the array \verb|ia|, column indices are stored in
the array \verb|ja|, and numerical values of corresponding elements are
stored in the array \verb|ar|. Then the statement \verb|s30| calls
the routine \verb|lpx_load_matrix|, which loads information from these
three arrays into the problem object.

Now all data have been entered into the problem object, and therefore
the statement \verb|s31| calls the routine \verb|lpx_simplex|, which is
a driver to the simplex method, in order to solve the LP problem. This
routine finds an optimal solution and stores all relevant information
back into the problem object.

The statement \verb|s32| obtains a computed value of the objective
function, and the statements \verb|s33|---\verb|s35| obtain computed
values of structural variables (columns), which correspond to the
optimal basic solution found by the solver.

The statement \verb|s36| prints the optimal solution to the standard
output. The printout may look like follows:

\begin{verbatim}
   Z = 733.333; x1 = 33.3333; x2 = 66.6667; x3 = 0
\end{verbatim}

Finally, the statement \verb|s37| calls the routine
\verb|lpx_delete_prob|, which frees all the memory allocated to the
problem object.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{API Routines}

This chapter describes GLPK API routines intended for using in
application programs.

\paragraph{Error handling}

If some GLPK API routine detects erroneous or incorrect data passed by
the application program, it sends appropriate diagnostic messages to the
standard output and then abnormally terminates the application program.
In most practical cases this allows to simplify programming avoiding
numerous checks of return codes. Thus, in order to prevent crashing the
application program should check all data, which are suspected to be
incorrect, before calling GLPK API routines.

Should note that this kind of error handling is used only in cases of
incorrect data passed by the application program. If, for example, the
application program calls some GLPK API routine to read data from an
input file and these data are incorrect, the GLPK API routine reports
about error in the usual way by means of return code.

\paragraph{Thread safety}

Currently GLPK API routines are non-reentrant and therefore cannot be
used in multi-thread programs.

\paragraph{Array indexing}

Normally all GLPK routines start array indexing from 1, not from 0
(except the specially stipulated cases). This means, for example, if
some vector $x$ of the length $n$ is passed as an array to some GLPK
routine, the latter expects vector components to be placed in locations
\verb|x[1]|, \verb|x[2]|, \dots, \verb|x[n]|, and the location
\verb|x[0]| normally is not used.

In order to avoid indexing errors it is most convenient and most
reliable to declare the array \verb|x| as follows:

\begin{verbatim}
   double x[1+n];
\end{verbatim}

\noindent
or to allocate it as follows:

\begin{verbatim}
   double *x;
   . . .
   x = calloc(1+n, sizeof(double));
\end{verbatim}

\noindent
In both cases one extra location \verb|x[0]| is reserved that allows
passing this array to GLPK routines in a usual way.

\paragraph{Using GLPK routines in C++ programs}

If you need to use GLPK routines in C++ programs, use the following
construction:

\begin{verbatim}
   extern "C" {
   #include "glpk.h"
   }
\end{verbatim}

\section{Problem object}

GLPK API routines deal with so called {\it problem objects}, which
are program objects of type \verb|LPX| intended to represent particular
LP and MIP problem instances.

The type \verb|LPX| is a data structure declared in the header file
\verb|glpk.h| as follows:

\begin{verbatim}
   typedef struct { ... } LPX;
\end{verbatim}

Problem objects (i.e. program objects of the \verb|LPX| type) are
allocated and managed internally by the GLPK API routines. The
application program should never use any members of the LPX structure
directly and should deal only with pointers to these objects (that is,
\verb|LPX| \verb|*| values).

Each problem object consists of four logical segments, which are:

$\bullet$ problem segment,

$\bullet$ basis segment,

$\bullet$ interior point segment,

$\bullet$ MIP segment, and

$\bullet$ control parameters and statistics segment.

\paragraph{Problem segment}

The {\it problem segment} contains original LP/MIP data, which
corresponds to the problem formulation (1.1)---(1.3) (see Section
\ref{seclp}, page \pageref{seclp}):

$\bullet$ rows (auxiliary variables),

$\bullet$ columns (structural variables),

$\bullet$ objective function, and

$\bullet$ constraint matrix.

Rows and columns have the same set of the following attributes:

$\bullet$ ordinal number,

$\bullet$ symbolic name (1 up to 255 arbitrary graphic characters),

$\bullet$ type (free, lower bound, upper bound, double bound, fixed),

$\bullet$ numerical values of lower and upper bounds,

$\bullet$ scale factor.

{\it Ordinal numbers} are intended for referencing rows and columns.
Row ordinal numbers are integers $1, 2, \dots, m$, and column ordinal
numbers are integers $1, 2, \dots, n$, where $m$ and $n$ are,
respectively, the current number of rows and columns in the problem
object.

{\it Symbolic names} are intended only for informational purposes. They
cannot be used for referencing rows and columns.

{\it Types and bounds} of rows (auxiliary variables) and columns
(structural variables) are explained above (see Section \ref{seclp},
page \pageref{seclp}).

{\it Scale factors} are used internally for scaling corresponding rows
and columns of the constraint matrix.

Information about the {\it objective function} includes numerical
values of objective coefficients and a flag, which defines the
optimization direction (i.e. minimization or maximization).

The {\it constraint matrix} is a $m \times n$ rectangular matrix built
of constraint coefficients $a_{ij}$, which defines the system of linear
constraints (1.2) (see Section \ref{seclp}, page \pageref{seclp}). This
matrix is stored in the problem object in both row-wise and column-wise
sparse formats.

Once the problem object has been created, the application program can
access and modify any components of the problem segment in arbitrary
order.

\paragraph{Basis segment}

The {\it basis segment} of the problem object keeps information related
to a current basic solution. This information includes:

$\bullet$ row and column statuses,

$\bullet$ basic solution statuses,

$\bullet$ factorization of the current basis matrix, and

$\bullet$ basic solution components.

The {\it row and column statuses} define which rows and columns are
basic and which are non-basic. These statuses may be assigned either by
the application program of by some API routines. Note that these
statuses are always defined independently on whether the corresponding
basis is valid or not.

The {\it basic solution statuses} include the {\it primal status} and
the {\it dual status}, which are set by the simplex-based solver once
the problem has been solved. The primal status shows whether a primal
basic solution is feasible, infeasible, or undefined. The dual status
shows the same for a dual basic solution.

The {\it factorization of the basis matrix} is some factorized form
(like LU-factorization) of the current basis matrix (defined by the
current row and column statuses). The factorization is used by the
simplex-based solver and kept when the solver terminates the search.
This feature allows efficiently reoptimizing the problem after some
modifications (for example, after changing some bounds or objective
coefficients). It also allows performing a post-optimal analysis (for
example, computing components of the simplex table, etc.).

The {\it basic solution components} include primal and dual values of
all auxiliary and structural variables for the most recently obtained
basic solution.

\paragraph{Interior point segment}

The {\it interior point segment} is automatically allocated after the
problem has been solved using the interior point solver. It contains
interior point solution components, which include the solution status,
and primal and dual values of all auxiliary and structural variables.

\paragraph{MIP segment}

The {\it MIP segment} is used only for MIP problems. This segment
includes:

$\bullet$ column kinds,

$\bullet$ MIP solution status, and

$\bullet$ MIP solution components.

The {\it column kinds} define which columns (i.e. structural variables)
are integer and which are continuous.

The {\it MIP solution status} is set by the MIP solver and shows whether
a MIP solution is integer optimal, integer feasible (non-optimal), or
undefined.

The {\it MIP solution components} are computed by the MIP solver and
includes primal values of all auxiliary and structural variables for the
most recently obtained MIP solution.

Note that in the case of MIP problem the basis segment corresponds to
an optimal solution of LP relaxation, which is also available to the
application program.

Currently the search tree is not kept in the MIP segment. Therefore if
the search has been terminated, it cannot be continued.

\paragraph{Control parameters and statistics segment}

This segment contains a fixed set of parameters, where each parameter
has the following three attributes:

$\bullet$ code,

$\bullet$ type, and

$\bullet$ current value.

The {\it parameter code} is intended for referencing a particular
parameter. All the parameter codes have symbolic names, which are macros
defined in the header file \verb|glpk.h|. Note that the parameter codes
are distinct positive integers.

The {\it parameter type} can be integer, real (floating-point), and
text (character string).

The {\it parameter value} is its current value kept in the problem
object. Initially (after the problem object has been created) all
parameters are assigned some default values.

Parameters are intended for several purposes. Some of them, which are
called {\it control parameters}, affect the behavior of API routines
(for example, the parameter \verb|LPX_K_ITLIM| limits maximal number of
simplex iterations available to the solver). Others, which are called
{\it statistics}, just represent some additional information about the
problem object (for example, the parameter \verb|LPX_K_ITCNT| shows how
many simplex iterations were performed for a particular problem object).

\newpage

\section{Problem creating and modifying routines}

\noindent \hrulefill

\subsection{{\tt lpx\_create\_prob} --- create problem object}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
LPX *lpx_create_prob(void);
\end{verbatim}

\paragraph{Description}
The routine \verb|lpx_create_prob| creates a new problem object, which
is ``empty'', i.e. has no rows and no columns.

\paragraph{Returns}
The routine returns a pointer to the created object, which should be
used in any subsequent operations on this object.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_prob\_name} --- assign (change) problem name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_prob_name(LPX *lp, char *name);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_prob_name| assigns a given symbolic
\verb|name| (1 up to 255 characters) to the specified problem object.

If the parameter \verb|name| is \verb|NULL| or empty string, the routine
erases an existing symbolic name of the problem object.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_obj\_name} --- assign (change) objective
function name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_obj_name(LPX *lp, char *name);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_obj_name| assigns a given symbolic
\verb|name| (1 up to 255 characters) to the objective function of the
specified problem object.

If the parameter \verb|name| is \verb|NULL| or empty string, the routine
erases an existing symbolic name of the objective function.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_obj\_dir} --- set (change) optimization
direction flag}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_obj_dir(LPX *lp, int dir);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_obj_dir| sets (changes) the optimization
direction flag (i.e. ``sense'' of the objective function) as specified
by the parameter \verb|dir|:

\begin{tabular}{@{}ll}
\verb|LPX_MIN| & minimization; \\
\verb|LPX_MAX| & maximization. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_add\_rows} --- add new rows to problem object}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_add_rows(LPX *lp, int nrs);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_add_rows| adds \verb|nrs| rows (constraints) to
the specified problem object. New rows are always added to the end of
the row list, so the ordinal numbers of existing rows are not changed.

Being added each new row is initially free (unbounded) and has empty
list of the constraint coefficients.

\paragraph{Returns}

The routine \verb|lpx_add_rows| returns the ordinal number of the first
new row added to the problem object.

\noindent \hrulefill

\subsection{{\tt lpx\_add\_cols} --- add new columns to problem object}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_add_cols(LPX *lp, int ncs);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_add_cols| adds \verb|ncs| columns (structural
variables) to the specified problem object. New columns are always added
to the end of the column list, so the ordinal numbers of existing
columns are not changed.

Being added each new column is initially fixed at zero and has empty
list of the constraint coefficients.

\paragraph{Returns}

The routine \verb|lpx_add_cols| returns the ordinal number of the first
new column added to the problem object.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_row\_name} --- assign (change) row name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_row_name(LPX *lp, int i, char *name);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_row_name| assigns a given symbolic
\verb|name| (1 up to 255 characters) to \verb|i|-th row (auxiliary
variable) of the specified problem object.

If the parameter \verb|name| is \verb|NULL| or empty string, the routine
erases an existing name of $i$-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_col\_name} --- assign (change) column name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_col_name(LPX *lp, int j, char *name);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_col_name| assigns a given symbolic
\verb|name| (1 up to 255 characters) to \verb|j|-th column (structural
variable) of the specified problem object.

If the parameter \verb|name| is \verb|NULL| or empty string, the routine
erases an existing name of $j$-th column.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_row\_bnds} --- set (change) row bounds}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_row_bnds(LPX *lp, int i, int type, double lb, double ub);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_row_bnds| sets (changes) the type and bounds
of \verb|i|-th row (auxiliary variable) of the specified problem object.

The parameters \verb|type|, \verb|lb|, and \verb|ub| specify the type,
lower bound, and upper bound, respectively, as follows:

\begin{center}
\begin{tabular}{cr@{}c@{}ll}
Type & \multicolumn{3}{c}{Bounds} & Comment \\
\hline
\verb|LPX_FR| & $-\infty <$ &$\ x\ $& $< +\infty$
   & Free (unbounded) variable \\
\verb|LPX_LO| & $lb \leq$ &$\ x\ $& $< +\infty$
   & Variable with lower bound \\
\verb|LPX_UP| & $-\infty <$ &$\ x\ $& $\leq ub$
   & Variable with upper bound \\
\verb|LPX_DB| & $lb \leq$ &$\ x\ $& $\leq ub$
   & Double-bounded variable \\
\verb|LPX_FX| & $lb =$ &$\ x\ $& $= ub$
   & Fixed variable \\
\end{tabular}
\end{center}

\noindent
where $x$ is the auxiliary variable associated with $i$-th row.

If the row has no lower bound, the parameter \verb|lb| is ignored. If
the row has no upper bound, the parameter \verb|ub| is ignored. If the
row is an equality constraint (i.e. the corresponding auxiliary variable
is of fixed type), only the parameter \verb|lb| is used while the
parameter \verb|ub| is ignored.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_col\_bnds} --- set (change) column bounds}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_col_bnds(LPX *lp, int j, int type, double lb, double ub);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_col_bnds| sets (changes) the type and bounds
of \verb|j|-th column (structural variable) of the specified problem
object.

The parameters \verb|type|, \verb|lb|, and \verb|ub| specify the type,
lower bound, and upper bound, respectively, as follows:

\begin{center}
\begin{tabular}{cr@{}c@{}ll}
Type & \multicolumn{3}{c}{Bounds} & Comment \\
\hline
\verb|LPX_FR| & $-\infty <$ &$\ x\ $& $< +\infty$
   & Free (unbounded) variable \\
\verb|LPX_LO| & $lb \leq$ &$\ x\ $& $< +\infty$
   & Variable with lower bound \\
\verb|LPX_UP| & $-\infty <$ &$\ x\ $& $\leq ub$
   & Variable with upper bound \\
\verb|LPX_DB| & $lb \leq$ &$\ x\ $& $\leq ub$
   & Double-bounded variable \\
\verb|LPX_FX| & $lb =$ &$\ x\ $& $= ub$
   & Fixed variable \\
\end{tabular}
\end{center}

\noindent
where $x$ is the structural variable associated with $j$-th column.

If the column has no lower bound, the parameter \verb|lb| is ignored.
If the column has no upper bound, the parameter \verb|ub| is ignored.
If the column is of fixed type, only the parameter \verb|lb| is used
while the parameter \verb|ub| is ignored.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_obj\_coef} --- set (change) objective
coefficient or constant term}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_obj_coef(LPX *lp, int j, double coef);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_obj_coef| sets (changes) the objective
coefficient at \verb|j|-th column (structural variable). A new value of
the objective coefficient is specified by the parameter \verb|coef|.

If the parameter \verb|j| is 0, the routine sets (changes) the constant
term (``shift'') of the objective function.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_mat\_row} --- set (replace) row of the
constraint matrix}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_mat_row(LPX *lp, int i, int len, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_mat_row| stores (replaces) the contents of
\verb|i|-th row of the constraint matrix of the specified problem
object.

Column indices and numerical values of new row elements must be placed
in locations \verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|,
\dots, \verb|val[len]|, respectively, where $0 \leq$ \verb|len| $\leq n$
is the new length of $i$-th row, $n$ is the current number of columns in
the problem object. Note that zero elements as well as elements with
identical column indices are not allowed.

If the parameter \verb|len| is 0, the parameters \verb|ind| and/or
\verb|val| can be specified as \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_mat\_col} --- set (replace) column of the
constraint matrix}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_mat_col(LPX *lp, int j, int len, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_mat_col| stores (replaces) the contents of
\verb|j|-th column of the constraint matrix of the specified problem
object.

Row indices and numerical values of new column elements must be placed
in locations \verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|,
\dots, \verb|val[len]|, respectively, where $0 \leq$ \verb|len| $\leq m$
is the new length of $j$-th column, $m$ is the current number of rows in
the problem object. Note that zero elements as well as elements with
identical row indices are not allowed.

If the parameter \verb|len| is 0, the parameters \verb|ind| and/or
\verb|val| can be specified as \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_load\_matrix} --- load (replace) the whole
constraint matrix}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_load_matrix(LPX *lp, int ne, int ia[], int ja[], double ar[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_load_matrix| loads the constraint matrix passed
in  the arrays \verb|ia|, \verb|ja|, and \verb|ar| into the specified
problem object. Before loading the current contents of the constraint
matrix is destroyed.

Constraint coefficients (elements of the constraint matrix) must be
specified as triplets (\verb|ia[k]|, \verb|ja[k]|, \verb|ar[k]|) for
$k=1,\dots,ne$, where \verb|ia[k]| is the row index, \verb|ja[k]| is
the column index, and \verb|ar[k]| is a numeric value of corresponding
constraint coefficient. The parameter \verb|ne| specifies the total
number of (non-zero) elements in the matrix to be loaded. Note that
coefficients with identical indices as well as zero coefficients are
not allowed.

If the parameter \verb|ne| is 0, the parameters \verb|ia|, \verb|ja|,
and/or \verb|ar| can be specified as \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_del\_rows} --- delete rows from problem object}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_del_rows(LPX *lp, int nrs, int num[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_del_rows| deletes specified rows from a problem
object, which the parameter \verb|lp| points to. Ordinal numbers of rows
to be deleted must be placed in locations \verb|num[1]|, \dots,
\verb|num[nrs]|, where ${\tt nrs}>0$.

Note that deleting rows involves changing ordinal numbers of other
rows remaining in the problem object. New ordinal numbers of the
remaining rows are assigned under the assumption that the original
order of rows is not changed. Let, for example, before deletion there
be five rows $a$, $b$, $c$, $d$, $e$ with ordinal numbers 1, 2, 3, 4, 5,
and let rows $b$ and $d$ have been deleted. Then after deletion the
remaining rows $a$, $c$, $e$ are assigned new oridinal numbers 1, 2, 3.

\noindent \hrulefill

\subsection{{\tt lpx\_del\_cols} --- delete columns from problem object}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_del_cols(LPX *lp, int ncs, int num[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_del_cols| deletes specified columns from a
problem object, which the parameter \verb|lp| points to. Ordinal numbers
of columns to be deleted must be placed in locations \verb|num[1]|,
\dots, \verb|num[ncs]|, where ${\tt ncs}>0$.

Note that deleting columns involves changing ordinal numbers of other
columns remaining in the problem object. New ordinal numbers of the
remaining columns are assigned under the assumption that the original
order of columns is not changed. Let, for example, before deletion there
be six columns $p$, $q$, $r$, $s$, $t$, $u$ with ordinal numbers 1, 2,
3, 4, 5, 6, and let columns $p$, $q$, $s$ have been deleted. Then after
deletion the remaining columns $r$, $t$, $u$ are assigned new ordinal
numbers 1, 2, 3.

\noindent \hrulefill

\subsection{{\tt lpx\_delete\_prob} --- delete problem object}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_delete_prob(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_delete_prob| deletes a problem object, which the
parameter \verb|lp| points to, freeing all the memory allocated to this
object.

\noindent \hrulefill

\newpage

\section{Problem retrieving routines}

\noindent \hrulefill

\subsection{{\tt lpx\_get\_prob\_name} --- retrieve problem name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
char *lpx_get_prob_name(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_prob_name| returns a pointer to an internal
buffer, which contains symbolic name of the problem. However, if the
problem has no assigned name, the routine returns \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_obj\_name} --- retrieve objective function
name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
char *lpx_get_obj_name(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_obj_name| returns a pointer to an internal
buffer, which contains symbolic name assigned to the objective
function. However, if the objective function has no assigned name, the
routine returns \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_obj\_dir} --- retrieve optimization
direction flag}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_obj_dir(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_obj_dir| returns the optimization direction
flag (i.e. ``sense'' of the objective function):

\begin{tabular}{@{}ll}
\verb|LPX_MIN| & minimization; \\
\verb|LPX_MAX| & maximization. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_num\_rows} --- retrieve number of rows}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_num_rows(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_num_rows| returns the current number of rows
in the specified problem object.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_num\_cols} --- retrieve number of columns}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_num_cols(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_num_cols| returns the current number of
columns the specified problem object.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_name} --- retrieve row name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
char *lpx_get_row_name(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_name| returns a pointer to an internal
buffer, which contains a symbolic name assigned to \verb|i|-th row.
However, if the row has no assigned name, the routine returns
\verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_name} --- retrieve column name}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
char *lpx_get_col_name(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_name| returns a pointer to an internal
buffer, which contains a symbolic name assigned to \verb|j|-th column.
However, if the column has no assigned name, the routine returns
\verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_type} --- retrieve row type}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_row_type(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_type| returns the type of \verb|i|-th
row, i.e. the type of corresponding auxiliary variable, as follows:

\begin{tabular}{@{}ll}
\verb|LPX_FR| & free (unbounded) variable; \\
\verb|LPX_LO| & variable with lower bound; \\
\verb|LPX_UP| & variable with upper bound; \\
\verb|LPX_DB| & double-bounded variable; \\
\verb|LPX_FX| & fixed variable. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_lb} --- retrieve row lower bound}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_row_lb(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_lb| returns the lower bound of
\verb|i|-th row, i.e. the lower bound of corresponding auxiliary
variable. However, if the row has no lower bound, the routine returns
zero.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_ub} --- retrieve row upper bound}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_row_ub(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_ub| returns the upper bound of
\verb|i|-th row, i.e. the upper bound of corresponding auxiliary
variable. However, if the row has no upper bound, the routine returns
zero.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_type} --- retrieve column type}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_col_type(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_type| returns the type of \verb|j|-th
column, i.e. the type of corresponding structural variable, as follows:

\begin{tabular}{@{}ll}
\verb|LPX_FR| & free (unbounded) variable; \\
\verb|LPX_LO| & variable with lower bound; \\
\verb|LPX_UP| & variable with upper bound; \\
\verb|LPX_DB| & double-bounded variable; \\
\verb|LPX_FX| & fixed variable. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_lb} --- retrieve column lower bound}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_col_lb(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_lb| returns the lower bound of
\verb|j|-th column, i.e. the lower bound of corresponding structural
variable. However, if the column has no lower bound, the routine returns
zero.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_ub} --- retrieve column upper bound}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_col_ub(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_ub| returns the upper bound of
\verb|j|-th column, i.e. the upper bound of corresponding structural
variable. However, if the column has no upper bound, the routine returns
zero.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_obj\_coef} --- retrieve objective coefficient
or constant term}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_obj_coef(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_obj_coef| returns the objective coefficient
at \verb|j|-th structural variable (column).

If the parameter \verb|j| is 0, the routine returns the constant term
(``shift'') of the objective function.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_num\_nz} --- retrieve number of constraint
coefficients}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_num_nz(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_num_nz| returns the number of non-zero
elements in the constraint matrix of the specified problem object.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_mat\_row} --- retrieve row of the constraint
matrix}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_mat_row(LPX *lp, int i, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_get_mat_row| scans (non-zero) elements of
\verb|i|-th row of the constraint matrix of the specified problem object
and stores their column indices and numeric values to locations
\verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|, \dots,
\verb|val[len]|, respectively, where $0\leq{\tt len}\leq n$ is the
number of elements in $i$-th row, $n$ is the number of columns.

The parameter \verb|ind| and/or \verb|val| can be specified as
\verb|NULL|, in which case corresponding information is not stored.

\paragraph{Returns}

The routine \verb|lpx_get_mat_row| returns the length \verb|len|, i.e.
the number of (non-zero) elements in \verb|i|-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_mat\_col} --- retrieve column of the
constraint matrix}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_mat_col(LPX *lp, int j, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_get_mat_col| scans (non-zero) elements of
\verb|j|-th column of the constraint matrix of the specified problem
object and stores their row indices and numeric values to locations
\verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|, \dots,
\verb|val[len]|, respectively, where $0\leq{\tt len}\leq m$ is the
number of elements in $j$-th column, $m$ is the number of rows.

The parameter \verb|ind| and/or \verb|val| can be specified as
\verb|NULL|, in which case corresponding information is not stored.

\paragraph{Returns}

The routine \verb|lpx_get_mat_col| returns the length \verb|len|, i.e.
the number of (non-zero) elements in \verb|j|-th column.

\noindent \hrulefill

\newpage

\section{Problem scaling routines}

\noindent \hrulefill

\subsection{{\tt lpx\_scale\_prob} --- scale problem data}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_scale_prob(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_scale_prob| performs scaling of problem data for
the specified problem object.

The purpose of scaling is to provide such scaling (diagonal) matrices
$R$ and $S$ that the scaled constraint matrix $A'=RAS$ has better
numerical properties than the original constraint matrix $A$.

Note that the scaling matrices $R$ and $S$ are used only by the solver.
On API level the scaling is invisible, since all data stored in the
problem object are non-scaled.

\noindent \hrulefill

\subsection{{\tt lpx\_unscale\_prob} --- unscale problem data}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_unscale_prob(LPX *lp);
\end{verbatim}

The routine \verb|lpx_unscale_prob| performs unscaling of problem data
for the specified problem object.

``Unscaling'' means replacing the current scaling matrices $R$ and $S$
by unity matrices that cancels the scaling effect.

\noindent \hrulefill

\newpage

\section{LP basis constructing routines}

\noindent \hrulefill

\subsection{{\tt lpx\_std\_basis} --- construct standard initial LP
basis}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_std_basis(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_std_basis| constructs the ``standard'' (trivial)
initial LP basis for the specified problem object.

In the ``standard'' LP basis all auxiliary variables (rows) are basic,
and all structural variables (columns) are non-basic (so the
corresponding basis matrix is unity).

\noindent \hrulefill

\subsection{{\tt lpx\_adv\_basis} --- construct advanced initial LP
basis}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_adv_basis(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_adv_basis| build an advanced initial LP basis for
the specified problem object.

In order to construct the advanced initial LP basis the routine does the
following:

1) includes in the basis all non-fixed auxiliary variables;

2) includes in the basis as many non-fixed structural variables as
possible keeping triangular form of the basis matrix;

3) includes in the basis appropriate (fixed) auxiliary variables to
complete the basis.

As a result the initial LP basis has as few fixed variables as possible
and the corresponding basis matrix is triangular.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_row\_stat} --- set (change) row status}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_row_stat(LPX *lp, int i, int stat);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_row_stat| sets (changes) the current status
of \verb|i|-th row (auxiliary variable) as specified by the parameter
\verb|stat|:

\begin{tabular}{@{}lp{127mm}@{}}
\verb|LPX_BS| & make the row basic (make the constraint inactive); \\
\verb|LPX_NL| & make the row non-basic (make the constraint active); \\
\verb|LPX_NU| & make the row non-basic and set it to the upper bound; if
   the row is not double-bounded, this status is equivalent to
   \verb|LPX_NL| (only in the case of this routine); \\
\verb|LPX_NF| & the same as \verb|LPX_NL| (only in the case of this
   routine); \\
\verb|LPX_NS| & the same as \verb|LPX_NL| (only in the case of this
   routine). \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_set\_col\_stat} --- set (change) column status}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_col_stat(LPX *lp, int j, int stat);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_col_stat sets| (changes) the current status
of \verb|j|-th column (structural variable) as specified by the
parameter \verb|stat|:

\begin{tabular}{@{}lp{127mm}@{}}
\verb|LPX_BS| & make the column basic; \\
\verb|LPX_NL| & make the column non-basic; \\
\verb|LPX_NU| & make the column non-basic and set it to the upper
   bound; if the column is not double-bounded, this status is equivalent
   to \verb|LPX_NL| (only in the case of this routine); \\
\verb|LPX_NF| & the same as \verb|LPX_NL| (only in the case of this
   routine); \\
\verb|LPX_NS| & the same as \verb|LPX_NL| (only in the case of this
   routine).
\end{tabular}

\medskip

\noindent \hrulefill

\newpage

\section{Simplex method routine}

\noindent \hrulefill

\subsection{{\tt lpx\_simplex} --- solve LP problem using the simplex
method}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_simplex(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_simplex| is an interface to an LP problem solver
based on the two-phase revised simplex method.

This routine obtains problem data from the problem object, which the
parameter \verb|lp| points to, calls the solver to solve the LP problem,
and stores an obtained basic solution and other relevant information
back into the problem object.

Since solving of large-scale problems may take a long time, the solver
reports some information about the current basic solution, which is sent
to the standard output. This information has the following format:

\begin{verbatim}
*nnn:   objval = xxx   infeas = yyy (ddd)
\end{verbatim}

\noindent
where: `\verb|nnn|' is the iteration number, `\verb|xxx|' is the current
value of the objective function (which is unscaled and has correct
sign), `\verb|yyy|' is the current sum of primal infeasibilities (which
is scaled and therefore may be used for visual estimating only),
`\verb|ddd|' is the current number of fixed basic variables. If the
asterisk `\verb|*|' precedes to `\verb|nnn|', the solver is searching
for an optimal solution (phase II), otherwise the solver is searching
for a primal feasible solution (phase I).

Note that the simplex solver currently implemented in GLPK is not
perfect. Although it has been successfully tested on a wide set of LP
problems, there are hard problems, which cannot be solved by the GLPK
simplex solver.

\paragraph{Using built-in LP presolver}

The simplex solver has {\it built-in LP presolver}, which is a
subprogram that transforms the original LP problem specified in the
problem object to an equivalent LP problem, which may be easier for
solving with the simplex method than the original one. This is attained
mainly due to reducing the problem size and improving its numeric
properties (for example, by removing some inactive constraints or by
fixing some non-basic variables). Once the transformed LP problem has
been solved, the presolver transforms its basic solution back to
a corresponding basic solution of the original problem.

Presolving is an optional feature of the routine \verb|lpx_simplex|,
and by default it is disabled. In order to enable the LP presolver the
user should set the control parameter \verb|LPX_K_PRESOL| on (see
Subsection \ref{subsecparlist}, page \pageref{subsecparlist}) before
calling the routine \verb|lpx_simplex|. As a rule presolving is useful
when the problem is solved for the first time, and it is not recommended
to use presolving when the problem should be re-optimized.

The presolving procedure is transparent to the API user in the sense
that all necessary processing is performed internally, and a basic
solution of the original problem recovered by the presolver is the same
as if it were computed directly, i.e. without presolving.

Note that the presolver is able to recover only optimal solutions. If
a computed solution is infeasible or non-optimal, the corresponding
solution of the original problem cannot be recovered and therefore
remains undefined. If the user needs to know a basic solution even if
it is infeasible or non-optimal, the presolver must be disabled.

\paragraph{Returns}

If the LP presolver is disabled (the flag \verb|LPX_K_PRESOL| is off),
the routine \verb|lpx_simplex| returns one of the following exit codes:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_OK|     & the LP problem has been successfully solved.
   (Note that, for example, if the problem has no feasible solution,
   this exit code is reported.) \\
\verb|LPX_E_FAULT|  & unable to start the search because either the
   problem has no rows/columns, or the initial basis is invalid, or the
   initial basis matrix is singular or ill-conditioned. \\
\verb|LPX_E_OBJLL|  & the search was prematurely terminated because the
   objective function being maximized has reached its lower limit and
   continues decreasing (the dual simplex only). \\
\verb|LPX_E_OBJUL|  & the search was prematurely terminated because the
   objective function being minimized has reached its upper limit and
   continues increasing (the dual simplex only). \\
\verb|LPX_E_ITLIM|  & the search was prematurely terminated because the
   simplex iterations limit has been exceeded. \\
\end{tabular}

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_TMLIM|  & the search was prematurely terminated because the
   time limit has been exceeded. \\
\verb|LPX_E_SING|   & the search was prematurely terminated due to the
   solver failure (the current basis matrix got singular or
   ill-conditioned). \\
\end{tabular}

If the LP presolver is enabled (the flag \verb|LPX_K_PRESOL| is on),
the routine \verb|lpx_simplex| returns one of the following exit codes:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_OK|     & optimal solution of the LP problem has been
   found. \\
\verb|LPX_E_FAULT|  & the LP problem has no rows and/or columns. \\
\verb|LPX_E_NOPFS|  & the LP problem has no primal feasible solution. \\
\verb|LPX_E_NODFS|  & the LP problem has no dual feasible solution. \\
\verb|LPX_E_ITLIM|  & same as above. \\
\verb|LPX_E_TMLIB|  & same as above. \\
\verb|LPX_E_SING|   & same as above. \\
\end{tabular}

\medskip

\noindent \hrulefill

\newpage

\section{Basic solution retrieving routines}

\noindent \hrulefill

\subsection{{\tt lpx\_get\_status} --- retrieve generic status of basic
solution}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_status(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_status| reports the generic status of the
current basic solution for the specified problem object as follows:

\begin{tabular}{@{}ll}
\verb|LPX_OPT|    & solution is optimal; \\
\verb|LPX_FEAS|   & solution is feasible; \\
\verb|LPX_INFEAS| & solution is infeasible; \\
\verb|LPX_NOFEAS| & problem has no feasible solution; \\
\verb|LPX_UNBND|  & problem has unbounded solution; \\
\verb|LPX_UNDEF|  & solution is undefined. \\
\end{tabular}

More detailed information about the status of basic solution can be
retrieved using the routines \verb|lpx_get_prim_stat| and
\verb|lpx_get_dual_stat|.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_prim\_stat} --- retrieve primal status of
basic solution}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_prim_stat(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_prim_stat| reports the primal status of the
basic solution for the specified problem object as follows:

\begin{tabular}{@{}ll}
\verb|LPX_P_UNDEF|  & primal solution is undefined; \\
\verb|LPX_P_FEAS|   & solution is primal feasible; \\
\verb|LPX_P_INFEAS| & solution is primal infeasible; \\
\verb|LPX_P_NOFEAS| & no primal feasible solution exists. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_dual\_stat} --- retrieve dual status of
basic solution}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_dual_stat(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_dual_stat| reports the dual status of the
basic solution for the specified problem object as follows:

\begin{tabular}{@{}ll}
\verb|LPX_D_UNDEF|  & dual solution is undefined; \\
\verb|LPX_D_FEAS|   & solution is dual feasible; \\
\verb|LPX_D_INFEAS| & solution is dual infeasible; \\
\verb|LPX_D_NOFEAS| & no dual feasible solution exists. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_obj\_val} --- retrieve objective value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_obj_val(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_obj_val| returns current value of the
objective function.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_stat} --- retrieve row status}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_row_stat(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_stat| returns current status assigned to
the auxiliary variable associated with \verb|i|-th row as follows:

\begin{tabular}{@{}ll}
\verb|LPX_BS| & basic variable; \\
\verb|LPX_NL| & non-basic variable on its lower bound; \\
\verb|LPX_NU| & non-basic variable on its upper bound; \\
\verb|LPX_NF| & non-basic free (unbounded) variable; \\
\verb|LPX_NS| & non-basic fixed variable. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_prim} --- retrieve row primal value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_row_prim(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_prim| returns primal value of the
auxiliary variable associated with \verb|i|-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_row\_dual} --- retrieve row dual value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_row_dual(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_row_dual| returns dual value (i.e. reduced
cost) of the auxiliary variable associated with \verb|i|-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_stat} --- retrieve column status}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_col_stat(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_stat| returns current status assigned to
the structural variable associated with \verb|j|-th column as follows:

\begin{tabular}{@{}ll}
\verb|LPX_BS| & basic variable; \\
\verb|LPX_NL| & non-basic variable on its lower bound; \\
\verb|LPX_NU| & non-basic variable on its upper bound; \\
\verb|LPX_NF| & non-basic free (unbounded) variable; \\
\verb|LPX_NS| & non-basic fixed variable. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_prim} --- retrieve column primal value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_col_prim(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_prim| returns primal value of the
structural variable associated with \verb|j|-th column.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_dual} --- retrieve column dual value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_col_dual(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_dual| returns dual value (i.e. reduced
cost) of the structural variable associated with \verb|j|-th column.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_ray\_info} --- retrieve non-basic variable
which causes unboundness}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_ray_info(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_ray_info| returns the number $k$ of some
non-basic variable $x_k$, which causes primal unboundness. If such a
variable cannot be identified, the routine returns zero.

If $1\leq k\leq m$, $x_k$ is $k$-th auxiliary variable, and if
$m+1\leq k\leq m+n$, $x_k$ is $(k-m)$-th structural variable, where $m$
is the number of rows, $n$ is the number of columns in the specified
problem object.

``Unboundness'' means that the variable $x_k$ is non-basic and able to
{\it infinitely} change in a feasible direction improving the objective
function.

\noindent \hrulefill

\subsection{{\tt lpx\_check\_kkt} --- check Karush-Kuhn-Tucker
conditions}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_check_kkt(LPX *lp, int scaled, LPXKKT *kkt);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_check_kkt| checks Karush-Kuhn-Tucker optimality
conditions for basic solution. It is assumed that both primal and dual
components of basic solution are valid.

If the parameter \verb|scaled| is zero, the optimality conditions are
checked for the original, unscaled LP problem. Otherwise, if the
parameter \verb|scaled| is non-zero, the routine checks the conditions
for an internally scaled LP problem.

The parameter \verb|kkt| is a pointer to the structure \verb|LPXKKT|,
to which the routine stores the results of checking. Members of this
structure are shown in the table below.

\begin{center}
\begin{tabular}{c|l|l}
Condition & Member & Comment \\
\hline
(KKT.PE) & \verb|pe_ae_max| &
         Largest absolute error \\
         & \verb|pe_ae_row| &
         Number of row with largest absolute error \\
         & \verb|pe_re_max| &
         Largest relative error \\
         & \verb|pe_re_row| &
         Number of row with largest relative error \\
         & \verb|pe_quality| &
         Quality of primal solution \\
\hline
(KKT.PB) & \verb|pb_ae_max| &
         Largest absolute error \\
         & \verb|pb_ae_ind| &
         Number of variable with largest absolute error \\
         & \verb|pb_re_max| &
         Largest relative error \\
         & \verb|pb_re_ind| &
         Number of variable with largest relative error \\
         & \verb|pb_quality| &
         Quality of primal feasibility \\
\hline
(KKT.DE) & \verb|de_ae_max| &
         Largest absolute error \\
         & \verb|de_ae_col| &
         Number of column with largest absolute error \\
         & \verb|de_re_max| &
         Largest relative error \\
         & \verb|de_re_col| &
         Number of column with largest relative error \\
         & \verb|de_quality| &
         Quality of dual solution \\
\hline
(KKT.DB) & \verb|db_ae_max| &
         Largest absolute error \\
         & \verb|db_ae_ind| &
         Number of variable with largest absolute error \\
         & \verb|db_re_max| &
         Largest relative error \\
         & \verb|db_re_ind| &
         Number of variable with largest relative error \\
         & \verb|db_quality| &
         Quality of dual feasibility \\
\end{tabular}
\end{center}

The routine performs all computations using only components of the
given LP problem and the current basic solution.

\paragraph{Background}

The first condition checked by the routine is:
$$x_R - A x_S = 0, \eqno{\rm (KKT.PE)}$$
where $x_R$ is the subvector of auxiliary variables (rows), $x_S$ is the
subvector of structural variables (columns), $A$ is the constraint
matrix. This condition expresses the requirement that all primal
variables must satisfy to the system of equality constraints of the
original LP problem. In case of exact arithmetic this condition would be
satisfied for any basic solution; however, in case of inexact
(floating-point) arithmetic, this condition shows how accurate the
primal basic solution is, that depends on accuracy of a representation
of the basis matrix used by the simplex method routines.

The second condition checked by the routine is:
$$l_k \leq x_k \leq u_k {\rm \ \ \ for\ all}\ k=1,\dots,m+n,
\eqno{\rm (KKT.PB)}$$
where $x_k$ is auxiliary ($1\leq k\leq m$) or structural
($m+1\leq k\leq m+n$) variable, $l_k$ and $u_k$ are, respectively,
lower and upper bounds of the variable $x_k$ (including cases of
infinite bounds). This condition expresses the requirement that all
primal variables must satisfy to bound constraints of the original LP
problem. Since in case of basic solution all non-basic variables are
placed on their bounds, actually the condition (KKT.PB) needs to be
checked for basic variables only. If the primal basic solution has
sufficient accuracy, this condition shows primal feasibility of the
solution.

The third condition checked by the routine is:
$${\rm grad}\;Z = c = (\tilde{A})^T \pi + d,$$
where $Z$ is the objective function, $c$ is the vector of objective
coefficients, $(\tilde{A})^T$ is a matrix transposed to the expanded
constraint matrix $\tilde{A} = (I|-A)$, $\pi$ is a vector of Lagrange
multipliers that correspond to equality constraints of the original LP
problem, $d$ is a vector of Lagrange multipliers that correspond to
bound constraints for all (auxiliary and structural) variables of the
original LP problem. Geometrically the third condition expresses the
requirement that the gradient of the objective function must belong to
the orthogonal complement of a linear subspace defined by the equality
and active bound constraints, i.e. that the gradient must be a linear
combination of normals to the constraint planes, where Lagrange
multipliers $\pi$ and $d$ are coefficients of that linear combination.

To eliminate the vector $\pi$ the third condition can be rewritten as:
$$
\left(\begin{array}{@{}c@{}}I \\ -A^T\end{array}\right) \pi =
\left(\begin{array}{@{}c@{}}d_R \\ d_S\end{array}\right) +
\left(\begin{array}{@{}c@{}}c_R \\ c_S\end{array}\right),
$$
or, equivalently:
$$
\begin{array}{r@{}c@{}c}
\pi + d_R&\ =\ &c_R, \\
-A^T\pi + d_S&\ =\ &c_S. \\
\end{array}
$$
Then substituting the vector $\pi$ from the first equation into the
second one we have:
$$A^T (d_R - c_R) + (d_S - c_S) = 0, \eqno{\rm (KKT.DE)}$$
where $d_R$ is the subvector of reduced costs of auxiliary variables
(rows), $d_S$ is the subvector of reduced costs of structural variables
(columns), $c_R$ and $c_S$ are subvectors of objective coefficients at,
respectively, auxiliary and structural variables, $A^T$ is a matrix
transposed to the constraint matrix of the original LP problem. In case
of exact arithmetic this condition would be satisfied for any basic
solution; however, in case of inexact (floating-point) arithmetic, this
condition shows how accurate the dual basic solution is, that depends on
accuracy of a representation of the basis matrix used by the simplex
method routines.

The last, fourth condition checked by the routine is:

\medskip

\begin{tabular}{r@{}c@{}ll}
&$\ d_k\ $& $=0,$&if $x_k$ is basic or free non-basic variable \\
$0\leq$&$\ d_k\ $&$<+\infty$&if $x_k$ is non-basic on its lower
(minimization) \\
&&&or upper (maximization) bound \\
$-\infty<$&$\ d_k\ $&$\leq 0$&if $x_k$ is non-basic on its upper
(minimization) \\
&&&or lower (maximization) bound \\
$-\infty<$&$\ d_k\ $&$<+\infty$&if $x_k$ is non-basic fixed variable \\
\end{tabular} \hfill (KKT.DB)

\medskip

\noindent
for all $k=1,\dots,m+n$, where $d_k$ is a reduced cost (Lagrange
multiplier) of auxiliary ($1\leq k\leq m$) or structural
($m+1\leq k\leq m+n$) variable $x_k$. Geometrically this condition
expresses the requirement that constraints of the original problem must
"hold" the point preventing its movement along the anti-gradient (in
case of minimization) or the gradient (in case of maximization) of the
objective function. Since in case of basic solution reduced costs of
all basic variables are placed on their (zero) bounds, actually the
condition (KKT.DB) needs to be checked for non-basic variables only.
If the dual basic solution has sufficient accuracy, this condition shows
dual feasibility of the solution.

Should note that the complete set of Karush-Kuhn-Tucker optimality
conditions also includes the fifth, so called complementary slackness
condition, which expresses the requirement that at least either a primal
variable $x_k$ or its dual counterpart $d_k$ must be on its bound for
all $k=1,\dots,m+n$. However, being always satisfied by definition for
any basic solution that condition is not checked by the routine.

To check the first condition (KKT.PE) the routine computes a vector of
residuals:
$$g = x_R - A x_S,$$
determines component of this vector that correspond to largest absolute
and relative errors:

\medskip

\hspace{40mm}
\verb|pe_ae_max| $\displaystyle{= \max_{1\leq i\leq m}|g_i|}$,

\medskip

\hspace{40mm}
\verb|pe_re_max| $\displaystyle{= \max_{1\leq i\leq m}
\frac{|g_i|}{1+|(x_R)_i|}}$,

\medskip

\noindent
and stores these quantities and corresponding row indices to the
structure \verb|LPXKKT|.

To check the second condition (KKT.PB) the routine computes a vector
of residuals:
$$
h_k = \left\{
\begin{array}{ll}
0,         & {\rm if}\ l_k \leq x_k \leq u_k \\
x_k - l_k, & {\rm if}\ x_k < l_k \\
x_k - u_k, & {\rm if}\ x_k > u_k \\
\end{array}
\right.
$$
for all $k=1,\dots,m+n$, determines components of this vector that
correspond to largest absolute and relative errors:

\medskip

\hspace{40mm}
\verb|pb_ae_max| $\displaystyle{= \max_{1\leq k \leq m+n}|h_k|}$,

\medskip

\hspace{40mm}
\verb|pb_re_max| $\displaystyle{= \max_{1\leq k \leq m+n}
\frac{|h_k|}{1+|x_k|}}$,

\medskip

\noindent
and stores these quantities and corresponding variable indices to the
structure \verb|LPXKKT|.

To check the third condition (KKT.DE) the routine computes a vector of
residuals:
$$u = A^T (d_R - c_R) + (d_S - c_S),$$
determines components of this vector that correspond to largest
absolute and relative errors:

\medskip

\hspace{40mm}
\verb|de_ae_max| $\displaystyle{= \max_{1\leq j\leq n}|u_j|}$,

\medskip

\hspace{40mm}
\verb|de_re_max| $\displaystyle{= \max_{1\leq j\leq n}
\frac{|u_j|}{1+|(d_S)_j - (c_S)_j|}}$,

\medskip

\noindent
and stores these quantities and corresponding column indices to the
structure \verb|LPXKKT|.

To check the fourth condition (KKT.DB) the routine computes a vector
of residuals:

$$
v_k = \left\{
\begin{array}{ll}
0,         & {\rm if}\ d_k\ {\rm has\ correct\ sign} \\
d_k,       & {\rm if}\ d_k\ {\rm has\ wrong\ sign} \\
\end{array}
\right.
$$
for all $k=1,\dots,m+n$, determines components of this vector that
correspond to largest absolute and relative errors:

\medskip

\hspace{40mm}
\verb|db_ae_max| $\displaystyle{= \max_{1\leq k\leq m+n}|v_k|}$,

\medskip

\hspace{40mm}
\verb|db_re_max| $\displaystyle{= \max_{1\leq k\leq m+n}
\frac{|v_k|}{1+|d_k - c_k|}}$,

\medskip

\noindent
and stores these quantities and corresponding variable indices to the
structure \verb|LPXKKT|.

Using the relative errors for all the four conditions the routine
\verb|lpx_check_kkt| also estimates a "quality" of the basic solution
from the standpoint of these conditions and stores corresponding
quality indicators to the structure \verb|LPXKKT|:

\verb|pe_quality| --- quality of primal solution;

\verb|pb_quality| --- quality of primal feasibility;

\verb|de_quality| --- quality of dual solution;

\verb|db_quality| --- quality of dual feasibility.

Each of these indicators is assigned to one of the following four
values:

\verb|'H'| means high quality,

\verb|'M'| means medium quality,

\verb|'L'| means low quality, or

\verb|'?'| means wrong or infeasible solution.

If all the indicators show high or medium quality (for an internally
scaled LP problem, i.e. when the parameter \verb|scaled| in a call to
the routine \verb|lpx_check_kkt| is non-zero), the user can be sure that
the obtained basic solution is quite accurate.

If some of the indicators show low quality, the solution can still be
considered as relevant, though an additional analysis is needed
depending on which indicator shows low quality.

If the indicator \verb|pe_quality| is assigned to \verb|'?'|, the
primal solution is wrong. If the indicator \verb|de_quality| is assigned
to \verb|'?'|, the dual solution is wrong.

If the indicator \verb|db_quality| is assigned to \verb|'?'| while
other indicators show a good quality, this means that the current
basic solution being primal feasible is not dual feasible. Similarly,
if the indicator \verb|pb_quality| is assigned to \verb|'?'| while
other indicators are not, this means that the current basic solution
being dual feasible is not primal feasible.

\noindent \hrulefill

\newpage

\section{LP basis and simplex table routines}

\noindent \hrulefill

\subsection{{\tt lpx\_warm\_up} --- ``warm up'' LP basis}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_warm_up(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_warm_up| ``warms up'' the LP basis for the
specified problem object using current statuses assigned to rows and
columns (i.e. to auxiliary and structural variables).

``Warming up'' includes reinverting (factorizing) the basis matrix (if
neccesary), computing primal and dual components as well as determining
primal and dual statuses of the basic solution.

\paragraph{Returns}

The routine \verb|lpx_warm_up| returns one of the following exit codes:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_OK| & the LP basis has been successfully ``warmed up''. \\
\verb|LPX_E_EMPTY|  & the problem has no rows and/or no columns. \\
\verb|LPX_E_BADB|   & the LP basis is invalid, because the number of
   basic variables is not the same as the number of rows. \\
\verb|LPX_E_SING|   & the basis matrix is numerically singular or
   ill-conditioned.
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_eval\_tab\_row} --- compute row of the simplex
table}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_eval_tab_row(LPX *lp, int k, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_eval_tab_row| computes a row of the current
simplex table for the basic variable, which is specified by the number
\verb|k|: if $1\leq k\leq m$, $x_k$ is $k$-th auxiliary variable; if
$m+1\leq k\leq m+n$, $x_k$ is $(k-m)$-th structural variable, where $m$
is the number of rows, $n$ is the number of columns. The current basis
must be available.

The routine stores column indices and numerical values of non-zero
elements of the computed row in sparse format to locations
\verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|, \dots,
\verb|val[len]|, respectively, where $0\leq{\tt len}\leq n$ is the
number of non-zeros returned on exit.

Element indices stored in the array \verb|ind| have the same sense as
the index \verb|k|, i.e. indices 1 to $m$ denote auxiliary variables
and indices $m+1$ to $m+n$ denote structural ones (all these variables
are non-basic by definition).

The computed row shows how the specified basic variable $x_k = (x_B)_i$
depends on non-basic variables:
$$
(x_B)_i = \alpha_{i1}(x_N)_1 + \alpha_{i2}(x_N)_2 + \dots +
\alpha_{in}(x_N)_n,
$$
where $\alpha_{ij}$ are elements of the simplex table row, $(x_N)_j$
are non-basic (auxiliary and structural) variables.

\paragraph{Returns}

The routine \verb|lpx_eval_tab_row| returns the number of non-zero
elements in the simplex table row stored in the arrays \verb|ind| and
\verb|val|.

\noindent \hrulefill

\subsection{{\tt lpx\_eval\_tab\_col} --- compute column of the simplex
table}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_eval_tab_col(LPX *lp, int k, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_eval_tab_col| computes a column of the current
simplex table for the non-basic variable, which is specified by the
number \verb|k|: if $1\leq k\leq m$, $x_k$ is $k$-th auxiliary variable;
if $m+1\leq k\leq m+n$, $x_k$ is $(k-m)$-th structural variable, where
$m$ is the number of rows, $n$ is the number of columns. The current
basis must be valid.

The routine stores row indices and numerical values of non-zero
elements of the computed column in sparse format to locations
\verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|, \dots,
\verb|val[len]|, respectively, where $0\leq{\tt len}\leq m$ is the
number of non-zeros returned on exit.

Element indices stored in the array \verb|ind| have the same sense as
the index \verb|k|, i.e. indices 1 to $m$ denote auxiliary variables and
indices $m+1$ to $m+n$ denote structural ones (all these variables are
basic by definition).

The computed column shows how the basic variables depend on the
specified non-basic variable $x_k = (x_N)_j$:
$$
\begin{array}{rl}
(x_B)_1 &= \dots + \alpha_{1j}(x_N)_j + \dots \\
(x_B)_2 &= \dots + \alpha_{2j}(x_N)_j + \dots \\
        &  \dots \dots \dots \\
(x_B)_m &= \dots + \alpha_{mj}(x_N)_j + \dots \\
\end{array}
$$
where $\alpha_{ij}$ are elements of the simplex table column, $(x_B)_i$
are basic (auxiliary and structural) variables.

\paragraph{Returns}

The routine \verb|lpx_eval_tab_col| returns the number of non-zero
elements in the simplex table column stored in the arrays \verb|ndx| and
\verb|val|.

\noindent \hrulefill

\subsection{{\tt lpx\_transform\_row} --- transform explicitly specified
row}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_transform_row(LPX *lp, int len, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_transform_row| performs the same operation as the
routine \verb|lpx_eval_tab_row|, except that the transformed row is
specified explicitly.

The explicitly specified row may be thought as a linear form:
$$x=a_1x_{m+1}+a_2x_{m+2}+\dots+a_nx_{m+n},\eqno(1)$$
where $x$ is an auxiliary variable for this row, $a_j$ are coefficients
of the linear form, $x_{m+j}$ are structural variables.

On entry column indices and numerical values of non-zero coefficients
$a_j$ of the transformed row should be placed in locations
\verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|, \dots,
\verb|val[len]|, where \verb|len| is number of non-zero coefficients.

This routine uses the system of equality constraints and the current
basis in order to express the auxiliary variable $x$ in (1) through the
current non-basic variables (as if the transformed row were added to
the problem object and the auxiliary variable $x$ were basic), i.e. the
resultant row has the form:
$$x=\alpha_1(x_N)_1+\alpha_2(x_N)_2+\dots+\alpha_n(x_N)_n,\eqno(2)$$
where $\alpha_j$ are influence coefficients, $(x_N)_j$ are non-basic
(auxiliary and structural) variables, $n$ is number of columns in the
specified problem object.

On exit the routine stores indices and numerical values of non-zero
coefficients $\alpha_j$ of the resultant row (2) in locations
\verb|ind[1]|, \dots, \verb|ind[len']| and \verb|val[1]|, \dots,
\verb|val[len']|, where $0\leq{\tt len'}\leq n$ is the number of
non-zero coefficients in the resultant row returned by the routine.
Note that indices of non-basic variables stored in the array \verb|ind|
correspond to original ordinal numbers of variables: indices 1 to $m$
mean auxiliary variables and indices $m+1$ to $m+n$ mean structural
ones.

\paragraph{Returns}

The routine \verb|lpx_transform_row| returns \verb|len'|, the number of
non-zero coefficients in the resultant row stored in the arrays
\verb|ind| and \verb|val|.

\noindent \hrulefill

\subsection{{\tt lpx\_transform\_col} --- transform explicitly specified
column}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_transform_col(LPX *lp, int len, int ind[], double val[]);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_transform_col| performs the same operation as the
routine \verb|lpx_eval_tab_col|, except that the transformed column is
specified explicitly.

The explicitly specified column may be thought as it were added to
the original system of equality constraints:
$$
\begin{array}{rl}
x_1 &= a_{11}x_{m+1}+\dots+a_{1n}x_{m+n}+a_1x \\
x_2 &= a_{21}x_{m+1}+\dots+a_{2n}x_{m+n}+a_2x \\
    &  \dots \dots \dots \\
x_m &= a_{m1}x_{m+1}+\dots+a_{mn}x_{m+n}+a_mx \\
\end{array} \eqno(1)
$$
where $x_i$ are auxiliary variables, $x_{m+j}$ are structural variables
(presented in the problem object), $x$ is a structural variable for the
explicitly specified column, $a_i$ are constraint coefficients for $x$.

On entry row indices and numerical values of non-zero coefficients
$a_i$ of the transformed column should be placed in locations
\verb|ind[1]|, \dots, \verb|ind[len]| and \verb|val[1]|, \dots,
\verb|val[len]|, where \verb|len| is number of non-zero coefficients.

This routine uses the system of equality constraints and the current
basis in order to express the current basic variables through the
structural variable $x$ in (1) (as if the transformed column were added
to the problem object and the variable $x$ were non-basic):
$$
\begin{array}{rl}
(x_B)_1 &= \dots + \alpha_{1}x \\
(x_B)_2 &= \dots + \alpha_{2}x \\
        &  \dots \dots \dots   \\
(x_B)_m &= \dots + \alpha_{m}x \\
\end{array} \eqno(2)
$$
where $\alpha_i$ are influence coefficients, $x_B$ are basic (auxiliary
and structural) variables, $m$ is number of rows in the specified
problem object.

On exit the routine stores indices and numerical values of non-zero
coefficients $\alpha_i$ of the resultant column (2) in locations
\verb|ind[1]|, \dots, \verb|ind[len']| and \verb|val[1]|, \dots,
\verb|val[len']|, where $0\leq{\tt len'}\leq m$ is the number of
non-zero coefficients in the resultant column returned by the routine.
Note that indices of basic variables stored in the array \verb|ind|
correspond to original ordinal numbers of variables, i.e. indices
1 to $m$ mean auxiliary variables, indices $m+1$ to $m+n$ mean
structural ones.

\paragraph{Returns}

The routine \verb|lpx_transform_col| returns \verb|len'|, the number of
non-zero coefficients in the resultant column stored in the arrays
\verb|ind| and \verb|val|.

\noindent \hrulefill

\subsection{{\tt lpx\_prim\_ratio\_test} --- perform primal ratio test}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_prim_ratio_test(LPX *lp, int len, int ind[], double val[],
      int how, double tol);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_prim_ratio_test| performs the primal ratio test
for an explicitly specified column of the simplex table.

The primal basic solution associated with an LP problem object, which
the parameter \verb|lp| points to, should be feasible. No components
of the LP problem object are changed by the routine.

The explicitly specified column of the simplex table shows how the
basic variables $x_B$ depend on some non-basic variable $y$ (which is
not necessarily presented in the problem object):
$$
\begin{array}{rl}
(x_B)_1 &= \dots + \alpha_{1}y \\
(x_B)_2 &= \dots + \alpha_{2}y \\
        &  \dots \dots \dots   \\
(x_B)_m &= \dots + \alpha_{m}y \\
\end{array} \eqno(1)
$$

The column (1) is specifed on entry to the routine using the sparse
format. Ordinal numbers of basic variables $(x_B)_i$ should be placed in
locations \verb|ind[1]|, \dots, \verb|ind[len]|, where ordinal number
1 to $m$ denote auxiliary variables, and ordinal numbers $m+1$ to $m+n$
denote structural variables. The corresponding non-zero coefficients
$\alpha_i$ should be placed in locations \verb|val[1]|, \dots,
\verb|val[len]|. The arrays \verb|ind| and \verb|val| are not changed by
the routine.

The parameter \verb|how| specifies in which direction the variable $y$
changes on entering the basis: $+1$ means increasing, $-1$ means
decreasing.

The parameter \verb|tol| is a relative tolerance (small positive number)
used by the routine to skip small $\alpha_i$ in the column (1).

The routine determines the ordinal number of some basic variable
(among specified in \verb|ind[1]|, \dots, \verb|ind[len]|), which
reaches its (lower or upper) bound first before any other basic
variables do and which therefore should leave the basis instead the
variable $y$ in order to keep primal feasibility, and returns it on
exit. If the choice cannot be made (i.e. if the adjacent basic solution
is primal unbounded due to $y$), the routine returns zero.

\paragraph{Note}

If the non-basic variable $y$ is presented in the LP problem object, the
column (1) can be computed using the routine \verb|lpx_eval_tab_col|.
Otherwise it can be computed using the routine \verb|lpx_transform_col|.

\paragraph{Returns}

The routine \verb|lpx_prim_ratio_test| returns the ordinal number of
some basic variable $(x_B)_i$, which should leave the basis instead the
variable $y$ in order to keep primal feasibility. If the adjacent basic
solution is primal unbounded and therefore the choice cannot be made,
the routine returns zero.

\noindent \hrulefill

\subsection{{\tt lpx\_dual\_ratio\_test} --- perform dual ratio test}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_dual_ratio_test(LPX *lp, int len, int ind[], double val[],
      int how, double tol);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_dual_ratio_test| performs the dual ratio test for
an explicitly specified row of the simplex table.

The dual basic solution associated with an LP problem object, which the
parameter \verb|lp| points to, should be feasible. No components of the
LP problem object are changed by the routine.

The explicitly specified row of the simplex table is a linear form,
which shows how some basic variable $y$ (not necessarily presented in
the problem object) depends on non-basic variables $x_N$:
$$y=\alpha_1(x_N)_1+\alpha_2(x_N)_2+\dots+\alpha_n(x_N)_n.\eqno(1)$$

The linear form (1) is specified on entry to the routine using the
sparse format. Ordinal numbers of non-basic variables $(x_N)_j$ should
be placed in locations \verb|ind[1]|, \dots, \verb|ind[len]|, where
ordinal numbers 1 to $m$ denote auxiliary variables, and ordinal numbers
$m+1$ to $m+n$ denote structural variables. The corresponding non-zero
coefficients $\alpha_j$ should be placed in locations \verb|val[1]|,
\dots, \verb|val[len]|. The arrays \verb|ind| and \verb|val| are not
changed by the routine.

The parameter \verb|how| specifies in which direction the variable $y$
changes on leaving the basis: $+1$ means increasing, $-1$ means
decreasing.

The parameter \verb|tol| is a relative tolerance (small positive number)
used by the routine to skip small $\alpha_j$ in the form (1).

The routine determines the ordinal number of some non-basic variable
(among specified in \verb|ind[1]|, \dots, \verb|ind[len]|), whose
reduced cost reaches its (zero) bound first before this happens for any
other non-basic variables and which therefore should enter the basis
instead the variable $y$ in order to keep dual feasibility, and returns
it on exit. If the choice cannot be made (i.e. if the adjacent basic
solution is dual unbounded due to $y$), the routine returns zero.

\paragraph{Note}

If the basic variable $y$ is presented in the LP problem object, the
row (1) can be computed using the routine \verb|lpx_eval_tab_row|.
Otherwise it can be computed using the routine \verb|lpx_transform_row|.

\paragraph{Returns}

The routine \verb|lpx_dual_ratio_test| returns the ordinal number of
some non-basic variable $(x_N)_j$, which should enter the basis instead
the variable $y$ in order to keep dual feasibility. If the adjacent
basic solution is dual unbounded and therefore the choice cannot be
made, the routine returns zero.

\noindent \hrulefill

\newpage

\section{Interior-point method routines}

\noindent \hrulefill

\subsection{{\tt lpx\_interior} --- solve LP problem using the
primal-dual interior-point method}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_interior(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_interior| is an interface to the LP problem
solver based on the primal-dual interior-point method.

This routine obtains problem data from the problem object, which the
parameter \verb|lp| points to, calls the solver to solve the LP problem,
and stores the found solution back in the problem object.

Interior-point methods (also known as barrier methods) are more modern
and more powerful numerical methods for large-scale linear programming.
They especially fit for very sparse LP problems and allow solving such
problems much faster than the simplex method.

Solving large LP problems may take a long time, so the routine
\verb|lpx_interior| displays information about every interior point
iteration\footnote{Unlike the simplex method the interior point method
usually needs 30---50 iterations (independently on the problem size) in
order to find an optimal solution.}. This information is sent to the
standard output and has the following format:

\begin{verbatim}
   nnn: F = fff; rpi = ppp; rdi = ddd; gap = ggg
\end{verbatim}

\noindent where \verb|nnn| is iteration number, \verb|fff| is the
current value of the objective function (in the case of maximization it
has wrong sign), \verb|ppp| is the current relative primal
infeasibility, \verb|ddd| is the current relative dual infeasibility,
and \verb|ggg| is the current primal-dual gap.

Should note that currently the GLPK interior-point solver does not
include many important features, in particular:

it is not able to process dense columns. Thus, if the constraint matrix
of the LP problem has dense columns, the solving process will be
inefficient;

it has no features against numerical instability. For some LP problems
premature termination may happen if the matrix $ADA^T$ becomes singular
or ill-conditioned;

it is not able to identify the optimal basis, which corresponds to the
found interior-point solution. 

\paragraph{Returns}

The routine \verb|lpx_interior| returns one of the following exit codes:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_OK|     & the LP problem has been successfully solved (to
   optimality). \\
\verb|LPX_E_FAULT|  & the solver can't start the search because either
   the problem has no rows and/or no columns, or some row has non-zero
   objective coefficient. \\
\verb|LPX_E_NOFEAS| & the problem has no feasible (primal or dual)
   solution. \\
\end{tabular}

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_NOCONV| & the search was prematurely terminated due to very
slow convergence or divergence. \\
\verb|LPX_E_ITLIM|  & the search was prematurely terminated because the
   simplex iterations limit has been exceeded. \\
\verb|LPX_E_INSTAB| & the search was prematurely terminated due to
numerical instability on solving Newtonian system. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_ipt\_status} --- retrieve status of
interior-point solution}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_ipt_status(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_ipt_status| reports the status of a solution
found by the interior-point solver as follows:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_T_UNDEF| & interior-point solution is undefined. \\
\verb|LPX_T_OPT|   & interior-point solution is optimal. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_ipt\_obj\_val} --- retrieve objective value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_ipt_obj_val(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_ipt_obj_val| returns value of the objective
function for interior-point solution.

\noindent \hrulefill

\subsection{{\tt lpx\_ipt\_row\_prim} --- retrieve row primal value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_ipt_row_prim(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_ipt_row_prim| returns primal value of the
auxiliary variable associated with \verb|i|-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_ipt\_row\_dual} --- retrieve row dual value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_ipt_row_dual(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_ipt_row_dual| returns dual value (i.e. reduced
cost) of the auxiliary variable associated with \verb|i|-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_ipt\_col\_prim} --- retrieve column primal value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_ipt_col_prim(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_ipt_col_prim| returns primal value of the
structural variable associated with \verb|j|-th column.

\noindent \hrulefill

\subsection{{\tt lpx\_ipt\_col\_dual} --- retrieve column dual value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_ipt_col_dual(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_ipt_col_dual| returns dual value (i.e. reduced
cost) of the structural variable associated with \verb|j|-th column.

\noindent \hrulefill

\newpage

\section{MIP routines}

\noindent \hrulefill

\subsection{{\tt lpx\_set\_class} --- set (change) problem class}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_class(LPX *lp, int klass);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_class| sets (changes) the class of the problem
object as specified by the parameter \verb|klass|:

\begin{tabular}{@{}ll}
\verb|LPX_LP|  & pure linear programming (LP) problem; \\
\verb|LPX_MIP| & mixed integer programming (MIP) problem. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_class} --- retrieve problem class}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_class(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_class| returns the class of the specified
problem object:

\begin{tabular}{@{}ll}
\verb|LPX_LP|  & pure linear programming (LP) problem; \\
\verb|LPX_MIP| & mixed integer programming (MIP) problem. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_set\_col\_kind} --- set (change) column kind}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_col_kind(LPX *lp, int j, int kind);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_col_kind| sets (changes) the kind of
\verb|j|-th column (structural variable) as specified by the parameter
\verb|kind|:

\begin{tabular}{@{}ll}
\verb|LPX_CV| & continuous variable; \\
\verb|LPX_IV| & integer variable. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_col\_kind} --- retrieve column kind}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_col_kind(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_col_kind| returns the kind of \verb|j|-th
column (structural variable) as follows:

\begin{tabular}{@{}ll}
\verb|LPX_CV| & continuous variable; \\
\verb|LPX_IV| & integer variable. \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_get\_num\_int} --- retrieve number of integer
columns}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_num_int(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_num_int| returns the number of columns
(structural variables), which are marked as integer.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_num\_bin} --- retrieve number of binary
columns}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_num_bin(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_num_bin| returns the number of columns
(structural variables), which are marked as integer and whose lower
bound is zero and upper bound is one.

\noindent \hrulefill

\subsection{{\tt lpx\_integer} --- solve MIP problem using the
branch-and-bound method}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_integer(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_integer| is an interface to the MIP problem solver
based on the branch-and-bound method.

This routine obtains problem data from the problem object, which the
parameter \verb|lp| points to, calls the solver to solve the MIP
problem, and stores an obtained solution and other relevant information
back in the problem object.

On entry to this routine the problem object must contain an optimal
basic solution for LP relaxation, which can be obtained by means of the
simplex-based solver (see the routine \verb|lpx_simplex|).

Solving many MIP problems may take a long time, so the solver reports
some information about best known solution, which is sent to the
standard output. This information has the following format:

\begin{verbatim}
+nnn: mip = xxx <rho> yyy gap (ppp; qqq)
\end{verbatim}

\noindent
where `\verb|nnn|' is the simplex iteration number; `\verb|xxx|' is a
value of the objective function for the best known integer feasible
solution (if no integer feasible solution has been found yet,
`\verb|xxx|' is the text `\verb|not found yet|'); `\verb|rho|' is the
string `\verb|>=|' (in case of minimization) or `\verb|<=|' (in case of
maximization); `\verb|yyy|' is a global bound for exact integer optimum
(i.e. the exact integer optimum is always in the range from `\verb|xxx|'
to `\verb|yyy|'); `\verb|gap|' is the relative mip gap, in percents,
computed as $gap=|xxx-yyy|/(|xxx|+DBL\_EPSILON)\cdot 100\%$ (if $gap$ is
greater than $999.9\%$, it is not printed); `\verb|ppp|' is the number
of subproblems in the active list, `\verb|qqq|' is the number of
subproblems which have been already fathomed and therefore removed from
the branch-and-bound search tree.

Note that the branch-and-bound solver implemented in GLPK uses easy
heuristics for branching and backtracking, and therefore it is not
perfect. Most probably this solver can be used for solving MIP problems
with one or two hundreds of integer variables. Hard or very large scale
MIP problems cannot be solved by this routine.

\paragraph{Returns}

The routine \verb|lpx_integer| returns one of the following exit codes:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_E_OK|     & the MIP problem has been successfully solved.
   (Note that, for example, if the problem has no integer feasible
   solution, this exit code is reported.) \\
\verb|LPX_E_FAULT|  & unable to start the search because either: \\
                    & the problem is not of MIP class, or \\
                    & the problem object doesn't contain optimal
                      solution for LP relaxation, or \\
                    & some integer variable has non-integer lower or
                      upper bound, or \\
                    & some row has non-zero objective coefficient. \\
\verb|LPX_E_ITLIM|  & the search was prematurely terminated because the
   simplex iterations limit has been exceeded. \\
\verb|LPX_E_TMLIM|  & the search was prematurely terminated because the
   time limit has been exceeded. \\
\verb|LPX_E_SING|   & the search was prematurely terminated due to the
   solver failure (the current basis matrix got singular or
   ill-conditioned). \\
\end{tabular}

\medskip

\noindent \hrulefill

\subsection{{\tt lpx\_mip\_status} --- retrieve status of MIP
solution}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_mip_status(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_mip_status| reports the status of a MIP solution
found by the branch-and-bound solver as follows:

\begin{tabular}{@{}p{25mm}p{114mm}@{}}
\verb|LPX_I_UNDEF| & MIP solution is undefined. \\
\verb|LPX_I_OPT|   & MIP solution is integer optimal. \\
\verb|LPX_I_FEAS|  & MIP solution is integer feasible, however its
   optimality has not been proven, perhaps due to premature termination
   of the search. \\
\verb|LPX_I_NOFEAS| & problem has no integer feasible solution (proven
   by the solver). \\
\end{tabular}

\medskip

\noindent \hrulefill

\pagebreak

\subsection{{\tt lpx\_mip\_obj\_val} --- retrieve objective value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_mip_obj_val(LPX *lp);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_mip_obj_val| returns value of the objective
function for MIP solution.

\noindent \hrulefill

\subsection{{\tt lpx\_mip\_row\_val} --- retrieve row value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_mip_row_val(LPX *lp, int i);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_mip_row_val| returns value of the auxiliary
variable associated with \verb|i|-th row.

\noindent \hrulefill

\subsection{{\tt lpx\_mip\_col\_val} --- retrieve column value}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_mip_col_val(LPX *lp, int j);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_mip_col_val| returns value of the structural
variable associated with \verb|j|-th column.

\noindent \hrulefill

\newpage

\section{Control parameters and statistics routines}

\noindent \hrulefill

\subsection{{\tt lpx\_reset\_parms} --- reset control parameters to
default values}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_reset_parms(LPX *lp);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_reset_parms| resets all control parameters
associated with a problem object, which the parameter \verb|lp| points
to, to their default values.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_int\_parm} --- set (change) integer control
parameter}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_int_parm(LPX *lp, int parm, int val);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_int_parm| sets (changes) the current value of
an integer control parameter \verb|parm|. The parameter \verb|val|
specifies a new value of the control parameter.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_int\_parm} --- query integer control
parameter}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_get_int_parm(LPX *lp, int parm);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_int_parm| returns the current value of an
integer control parameter \verb|parm|.

\noindent \hrulefill

\subsection{{\tt lpx\_set\_real\_parm} --- set (change) real control
parameter}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
void lpx_set_real_parm(LPX *lp, int parm, double val);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_set_real_parm| sets (changes) the current value
of a real (floating point) control parameter \verb|parm|. The parameter
\verb|val| specifies a new value of the control parameter.

\noindent \hrulefill

\subsection{{\tt lpx\_get\_real\_parm} --- query real control parameter}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
double lpx_get_real_parm(LPX *lp, int parm);
\end{verbatim}

\paragraph{Returns}

The routine \verb|lpx_get_real_parm| returns the current value of a real
(floating point) control parameter \verb|parm|.

\noindent \hrulefill

\subsection{Parameter list}
\label{subsecparlist}

This subsection describes all control parameters currently implemented
in the package. Symbolic names of control parameters (which are macros
defined in the header file \verb|glpk.h|) are given on the left. Types,
default values, and descriptions are given on the right.

\bigskip

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MSGLEV|
   & type: integer, default: 3 \\
   & Level of messages output by solver routines: \\
   & 0 --- no output \\
   & 1 --- error messages only \\
   & 2 --- normal output \\
   & 3 --- full output (includes informational messages) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_SCALE|
   & type: integer, default: 1 \\
   & Scaling option: \\
   & 0 --- no scaling \\
   & 1 --- equilibration scaling \\
   & 2 --- geometric mean scaling \\
   & 3 --- geometric mean scaling, then equilibration scaling \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_DUAL|
   & type: integer, default: 0 \\
   & Dual simplex option: \\
   & 0 --- do not use the dual simplex \\
   & 1 --- if initial basic solution is dual feasible, use the dual
           simplex \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_PRICE|
   & type: integer, default: 1 \\
   & Pricing option (for both primal and dual simplex): \\
   & 0 --- textbook pricing \\
   & 1 --- steepest edge pricing \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_RELAX|
   & type: real, default: 0.07 \\
   & Relaxation parameter used in the ratio test. If it is zero, the
     textbook ratio test is used. If it is non-zero (should be
     positive), Harris' two-pass ratio test is used. In the latter case
     on the first pass of the ratio test basic variables (in the case of
     primal simplex) or reduced costs of non-basic variables (in the
     case of dual simplex) are allowed to slightly violate their bounds,
     but not more than (\verb|RELAX| $\cdot$ \verb|TOLBND|) or
     (\verb|RELAX| $\cdot$ \verb|TOLDJ|) (thus, \verb|RELAX| is a
     percentage of \verb|TOLBND| or \verb|TOLDJ|). \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_TOLBND|
   & type: real, default: $10^{-7}$ \\
   & Relative tolerance used to check if the current basic solution is
     primal feasible. (Do not change this parameter without detailed
     understanding its purpose.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_TOLDJ|
   & type: real, default: $10^{-7}$ \\
   & Absolute tolerance used to check if the current basic solution
     is dual feasible. (Do not change this parameter without detailed
     understanding its purpose.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_TOLPIV|
   & type: real, default: $10^{-9}$ \\
   & Relative tolerance used to choose eligible pivotal elements of
     the simplex table. (Do not change this parameter without detailed
     understanding its purpose.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_ROUND|
   & type: integer, default: 0 \\
   & Solution rounding option: \\
   & 0 --- report all primal and dual values ``as is'' \\
   & 1 --- replace tiny primal and dual values by exact zero \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_OBJLL|
   & type: real, default: \verb|-DBL_MAX| \\
   & Lower limit of the objective function. If on the phase II the
     objective function reaches this limit and continues decreasing, the
     solver stops the search. (Used in the dual simplex only.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_OBJUL|
   & type: real, default: \verb|+DBL_MAX| \\
   & Upper limit of the objective function. If on the phase II the
     objective function reaches this limit and continues increasing, the
     solver stops the search. (Used in the dual simplex only.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_ITLIM|
   & type: integer, default: $-1$ \\
   & Simplex iterations limit. If this value is positive, it is
     decreased by one each time when one simplex iteration has been
     performed, and reaching zero value signals the solver to stop the
     search. Negative value means no iterations limit. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_ITCNT|
   & type: integer, initial: 0 \\
   & Simplex iterations count. This count is increased by one each time
     when one simplex iteration has been performed. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_TMLIM|
   & type: real, default: $-1.0$ \\
   & Searching time limit, in seconds. If this value is positive, it is
     decreased each time when one simplex iteration has been performed
     by the amount of time spent for the iteration, and reaching zero
     value signals the solver to stop the search. Negative value means
     no time limit. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_OUTFRQ|
   & type: integer, default: 200 \\
   & Output frequency, in iterations. This parameter specifies how
     frequently the solver sends information about the solution to the
     standard output. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_OUTDLY|
   & type: real, default: 0.0 \\
   & Output delay, in seconds. This parameter specifies how long the
     solver should delay sending information about the solution to the
     standard output. Non-positive value means no delay. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_BRANCH|
   & type: integer, default: 2 \\
   & Branching heuristic option (for MIP only): \\
   & 0 --- branch on the first variable \\
   & 1 --- branch on the last variable \\
   & 2 --- branch using a heuristic by Driebeck and Tomlin \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_BTRACK|
   & type: integer, default: 2 \\
   & Backtracking heuristic option (for MIP only): \\
   & 0 --- depth first search \\
   & 1 --- breadth first search \\
   & 2 --- backtrack using the best projection heuristic \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_TOLINT|
   & type: real, default: $10^{-5}$ \\
   & Absolute tolerance used to check if the current basic solution
     is integer feasible. (Do not change this parameter without detailed
     understanding its purpose.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_TOLOBJ|
   & type: real, default: $10^{-7}$ \\
   & Relative tolerance used to check if the value of the objective
     function is not better than in the best known integer feasible
     solution. (Do not change this parameter without detailed
     understanding its purpose.) \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MPSINFO|
   & type: int, default: 1 \\
   & If this flag is set, the routine \verb|lpx_write_mps| writes
     several comment cards, which contains some information about the
     problem. Otherwise the routine writes no comment cards. This flag
     also affects the routine \verb|lpx_write_bas|.\\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MPSOBJ|
   & type: int, default: 2 \\
   & This parameter tells the routine \verb|lpx_write_mps| how to
     output the objective function row: \\
   & 0 --- never output objective function row \\
   & 1 --- always output objective function row \\
   & 2 --- output objective function row if the problem has no free
           rows \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MPSORIG|
   & type: int, default: 0 \\
   & If this flag is set, the routine \verb|lpx_write_mps| uses the
     original symbolic names of rows and columns. Otherwise the routine
     generates plain names using ordinal numbers of rows and columns.
     This flag also affects the routines \verb|lpx_read_bas| and
     \verb|lpx_write_bas|.\\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MPSWIDE|
   & type: int, default: 1 \\
   & If this flag is set, the routine \verb|lpx_write_mps| uses all
     data fields. Otherwise the routine keeps fields 5 and 6 empty. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MPSFREE|
   & type: int, default: 0 \\
   & If this flag is set, the routine \verb|lpx_write_mps| omits column
     and vector names every time when possible (free style). Otherwise
     the routine never omits these names (pedantic style). \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_MPSSKIP|
   & type: int, default: 0 \\
   & If this flag is set, the routine \verb|lpx_write_mps| skips empty
     columns (i.e. which has no constraint coefficients). Otherwise the
     routine outputs all columns. \\
\end{tabular}

\noindent\begin{tabular}{@{}p{28mm}p{117mm}@{}}
\verb|LPX_K_PRESOL|
   & type: int, default: 0 \\
   & If this flag is set, the routine \verb|lpx_simplex| solves the
     problem using the built-in LP presolver. Otherwise the LP presolver
     is not used. \\
\end{tabular}

\newpage

\section{Utility routines}

\noindent \hrulefill

\subsection{{\tt lpx\_read\_mps} --- read problem data in fixed MPS
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
LPX *lpx_read_mps(char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_read_mps| reads LP/MIP problem data in fixed MPS
format from an input text file whose name is the character string
\verb|fname|. (The MPS format is described in Appendix \ref{champs},
page \pageref{champs}.)

Behavior of the routine \verb|lpx_read_mps| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no error occurred, the routine returns a pointer to the created
problem object. Otherwise the routine returns \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_write\_mps} --- write problem data in fixed MPS
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_write_mps(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_write_mps| writes problem data in fixed MPS format
to an output text file whose name is the character string \verb|fname|.
(The MPS format is described in Appendix \ref{champs}, page
\pageref{champs}.)

Behavior of the routine \verb|lpx_write_mps| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_read\_bas} --- read LP basis in fixed MPS format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_read_bas(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_read_bas| reads LP basis in fixed MPS format from
an input text file whose name is the character string \verb|fname|.
(About this feature of the MPS format see Section \ref{secbas}, page
\pageref{secbas}.)

Behavior of the routine \verb|lpx_read_bas| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_write\_bas} --- write LP basis in fixed MPS
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_write_bas(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_write_bas| writes current LP basis in fixed MPS
format to an output text file whose name is the character string
\verb|fname|. (About this feature of the MPS format see Section
\ref{secbas}, page \pageref{secbas}.)

Behavior of the routine \verb|lpx_write_bas| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_read\_freemps} --- read problem data in free MPS
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
LPX *lpx_read_freemps(char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_read_freemps| reads LP/MIP problem data in free
MPS format from an input text file whose name is the character string
\verb|fname|. (The MPS format is described in Appendix \ref{champs},
page \pageref{champs}.)

Behavior of the routine \verb|lpx_read_freemps| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no error occurred, the routine returns a pointer to the created
problem object. Otherwise the routine returns \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_write\_freemps} --- write problem data in free MPS
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_write_freemps(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_write_freemps| writes problem data in fixed MPS
format to an output text file whose name is the character string
\verb|fname|. (The MPS format is described in Appendix \ref{champs},
page \pageref{champs}.)

Behavior of the routine \verb|lpx_write_freemps| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_read\_cpxlp} --- read problem data in CPLEX LP
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
LPX *lpx_read_cpxlp(char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_read_cpxlp| reads LP/MIP problem data in CPLEX LP
format from an input text file whose name is the character string
\verb|fname|. (The CPLEX LP format is described in Appendix
\ref{chacplex}, page \pageref{chacplex}.)

\paragraph{Returns}

Behavior of the routine \verb|lpx_read_cpxlp| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no error occurred, the routine returns a pointer to the created
problem object. Otherwise the routine returns \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_write\_cpxlp} --- write problem data in CPLEX LP
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_write_cpxlp(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_write_cpxlp| writes problem data in CPLEX LP
format to an output text file whose name is the character string
\verb|fname|. (The CPLEX LP format is described in Appendix
\ref{chacplex}, page \pageref{chacplex}.)

Behavior of the routine \verb|lpx_write_cpxlp| depends on some control
parameters (see Subsection \ref{subsecparlist}, page
\pageref{subsecparlist}.)

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_read\_model} --- read model written in GNU
MathProg modeling language}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
LPX *lpx_read_model(char *model, char *data, char *output);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_read_model| reads and translates LP/MIP model
(problem) written in the GNU MathProg modeling language.\footnote{The
GNU MathProg modeling language is a subset of the AMPL language.}

The character string \verb|model| specifies name of input text file,
which contains model section and, optionally, data section. This
parameter cannot be \verb|NULL|.

The character string \verb|data| specifies name of input text file,
which contains data section. This parameter can be \verb|NULL|. (If the
data file is specified and the model file also contains data section,
that section is ignored and data section from the data file is used.)

The character string \verb|output| specifies name of output text file,
to which the output produced by display statements is written. If the
parameter output is \verb|NULL|, the display output is sent to stdout
via the routine \verb|print|.

The routine \verb|lpx_read_model| is an interface to the model
translator, which is a program that parses model description and
translates it to some internal data structures.

For detailed description of the modeling language see the document
``GLPK: Modeling Language GNU MathProg'' included in the GLPK
distribution.

\paragraph{Returns}

If no errors occurred, the routine returns a pointer to the created
problem object. Otherwise the routine sends diagnostics to the standard
output and returns \verb|NULL|.

\noindent \hrulefill

\subsection{{\tt lpx\_print\_prob} --- write problem data in plain text
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_print_prob(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_print_prob| writes data from a problem object,
which the parameter \verb|lp| points to, to an output text file, whose
name is the character string \verb|fname|, in plain text format.

Information reported by the routine \verb|lpx_print_prob| is intended
mainly for visual analysis.

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_print\_sol} --- write basic solution in printable
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_print_sol(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_print_sol writes| the current basic solution of
an LP problem, which is specified by the pointer \verb|lp|, to a text
file, whose name is the character string \verb|fname|, in printable
format.

Information reported by the routine \verb|lpx_print_sol| is intended
mainly for visual analysis.

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_print\_sens\_bnds} --- write bounds sensitivity
information}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_print_sens_bnds(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_print_sens_bnds| writes the bounds for objective
coefficients, right-hand-sides of constraints, and variable bounds
for which the current optimal basic solution remains optimal (for LP
only).

The LP is given by the pointer \verb|lp|, and the output is written to
the file specified by \verb|fname|.  The current contents of the file
will be overwritten.

Information reported by the routine \verb|lpx_print_sens_bnds| is
intended mainly for visual analysis.

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_print\_ips} --- write interior point solution in
printable format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_print_ips(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_print_ips| writes the current interior point
solution  of an LP problem, which the parameter \verb|lp| points to, to
a text file, whose name is the character string \verb|fname|, in
printable format.

Information reported by the routine \verb|lpx_print_ips| is intended
mainly  for visual analysis.

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

\subsection{{\tt lpx\_print\_mip} --- write MIP solution in printable
format}

\paragraph{Synopsis}

\begin{verbatim}
#include "glpk.h"
int lpx_print_mip(LPX *lp, char *fname);
\end{verbatim}

\paragraph{Description}

The routine \verb|lpx_print_mip| writes a best known integer solution
of a MIP problem, which is specified by the pointer \verb|lp|, to a text
file, whose name is the character string \verb|fname|, in printable
format.

Information reported by the routine \verb|lpx_print_mip| is intended
mainly for visual analysis.

\paragraph{Returns}

If no errors occurred, the routine returns zero. Otherwise the routine
prints an error message and returns non-zero.

\noindent \hrulefill

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\chapter{Installing GLPK on Your Computer}

\section{Obtaining GLPK distribution file}

The distrubution file for the most recent version of the GLPK package
can be downloaded from $<$\verb|ftp://ftp.gnu.org/gnu/glpk/|$>$ or from
some mirror GNU ftp sites; for details see
$<$\verb|http://www.gnu.org/order/ftp.html|$>$.

\section{Unpacking the distribution file}

The GLPK package (like all other GNU software) is distributed in the
form of packed archive. This is one file named \verb|glpk-x.y.tar.gz|,
where {\it x} is the major version number and {\it y} is the minor
version number.

In order to prepare the distribution for installation you should:

1. Copy the GLPK distribution file to some subdirectory.

2. Enter the command \verb|gzip -d glpk-x.y.tar.gz| in order to unpack
the distribution file. After unpacking the name of the distribution file
will be automatically changed to \verb|glpk-x.y.tar|.

3. Enter the command \verb|tar -x < glpk-x.y.tar| in order to unarchive
the distribution. After this operation the subdirectory \verb|glpk-x.y|,
which is the GLPK distribution, will be automatically created.

\section{Configuring the package}

After you have unpacked and unarchived GLPK distribution you should
configure the package, i.e. automatically tune it for your computer
(platform).

Normally, you should just \verb|cd| to the subdirectory
\verb|glpk-x.y| and enter the command \verb|./configure|. If you are
using \verb|csh| on an old version of System V, you might need to type
\verb|sh configure| instead to prevent \verb|csh| from trying execute
\verb|configure| itself.

The \verb|configure| shell script attempts to guess correct values for
various system-dependent variables used during compilation, and creates
\verb|Makefile|. It also creates a file \verb|config.status| that you
can run in the future to recreate the current configuration.

Running \verb|configure| takes about a few minutes. While it is running,
it displays some informational messages that tell you what it is doing.
If you don't want to see these messages, run \verb|configure| with its
standard output redirected to \verb|dev/null|; for example,
\verb|./configure >/dev/null|.

\section{Compiling and checking the package}

Normally, in order to compile the package you should just enter the
command \verb|make|. This command reads \verb|Makefile| generated by
\verb|configure| and automatically performs all necessary job.

The result of compilation is:

$\bullet$ the file \verb|libglpk.a|, which is a library archive that
contains object code for all GLPK routines; and

$\bullet$ the program \verb|glpsol|, which is a stand-alone LP/MIP
solver.

If you want, you can override the \verb|make| variables \verb|CFLAGS|
and \verb|LDFLAGS| like this:

\verb|make CFLAGS=-O2 LDFLAGS=-s|

To compile the package in a different directory from the one containing
the source code, you must use a version of \verb|make| that supports
\verb|VPATH| variable, such as GNU make. \verb|cd| to the directory
where you want the object files and executables to go and run the
\verb|configure| script. \verb|configure| automatically checks for the
source code in the directory that \verb|configure| is in and in
`\verb|..|'. If for some reason \verb|configure| is not in the source
code directory that you are configuring, then it will report that it
can't find the source code. In that case, run \verb|configure| with the
option \verb|--srcdir=DIR|, where \verb|DIR| is the directory that
contains the source code.

On systems that require unusual options for compilation or linking the
package's \verb|configure| script does not know about, you can give
\verb|configure| initial values for variables by setting them in the
environment. In Bourne-compatible shells you can do that on the command
line like this:

\verb|CC='gcc -traditional' LIBS=-lposix ./configure|

Here are the \verb|make| variables that you might want to override with
environment variables when running \verb|configure|.

For these variables, any value given in the environment overrides the
value that \verb|configure| would choose:

$\bullet$ variable \verb|CC|: C compiler program. The default is
\verb|cc|.

$\bullet$ variable \verb|INSTALL|: program to use to install files. The
default value is \verb|install| if you have it, otherwise \verb|cp|.

For these variables, any value given in the environment is added to the
value that \verb|configure| chooses:

$\bullet$ variable DEFS: configuration options, in the form
`\verb|-Dfoo -Dbar| \dots'.

$\bullet$ variable LIBS: libraries to link with, in the form
`\verb|-lfoo -lbar| \dots'.

In order to check the package (running some tests included in the
distribution) you can just enter the command \verb|make check|.

\section{Installing the package}

Normally, in order to install the GLPK package (i.e. copy GLPK library,
header files, and the solver to the system places) you should just enter
the command \verb|make install| (note that you should be the root user
or a superuser).

By default, \verb|make install| will install the package's files in
the subdirectories \verb|usr/local/bin|, \verb|usr/local/lib|, etc. You
can specify an installation prefix other than \verb|/usr/local| by
giving \verb|configure| the option \verb|--prefix=PATH|. Alternately,
you can do so by consistently giving a value for the \verb|prefix|
variable when you run \verb|make|, e.g.

\verb|make prefix=/usr/gnu|

\verb|make prefix=/usr/gnu install|

After installing you can remove the program binaries and object files
from the source directory by typing \verb|make clean|. To remove all
files that \verb|configure| created (\verb|Makefile|,
\verb|config.status|, etc.), just type \verb|make distclean|.

The file \verb|configure.in| is used to create \verb|configure| by a
program called \verb|autoconf|. You only need it if you want to remake
\verb|configure| using a newer version of \verb|autoconf|.

\section{Uninstalling the package}

In order to uninstall the GLPK package (i.e. delete all GLPK files from
the system places) you can enter the command \verb|make uninstall|.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{MPS Format}
\label{champs}

\section{Fixed MPS Format}
\label{secmps}

The MPS format\footnote{The MPS format was developed in 1960's by IBM
as input format for their mathematical programming system MPS/360.
Today the MPS format is a most widely used format understood by most
mathematical programming packages. This appendix describes only the
features of the MPS format, which are implemented in the GLPK package.}
is intended for coding LP/MIP problem data. This format assumes the
formulation of LP/MIP problem (1.1)---(1.3) (see Section \ref{seclp},
page \pageref{seclp}).

{\it MPS file} is a text file, which contains two types of
cards\footnote{In 1960's MPS file was a deck of 80-column punched cards,
so the author decided to keep the word ``card'', which may be understood
as ``line of text file''.}: indicator cards and data cards.

Indicator cards determine a kind of succeeding data. Each indicator card
has one word in uppercase letters beginning in column 1.

Data cards contain problem data. Each data card is divided into six
fixed fields:

\begin{center}
\begin{tabular}{lcccccc}
& Field 1 & Field 2 & Field 3 & Field 4 & Field 5 & Feld 6 \\
\hline
Columns & 2---3 & 5---12 & 15---22 & 25---36 & 40---47 & 50---61 \\
Contents & Code & Name & Name & Number & Name & Number \\
\end{tabular}
\end{center}

On a particular data card some fields may be optional.

Names are used to identify rows, columns, and some vectors (see below).

Aligning the indicator code in the field 1 to the left margin is
optional.

All names specified in the fields 2, 3, and 5 should contain from 1 up
to 8 arbitrary characters (except control characters). If a name is
placed in the field 3 or 5, its first character should not be the dollar
sign `\verb|$|'. If a name contains spaces, the spaces are ignored.

All numerical values in the fields 4 and 6 should be coded in the form
$sxx$\verb|E|$syy$, where $s$ is the plus `\verb|+|' or the minus
`\verb|-|' sign, $xx$ is a real number with optional decimal point,
$yy$ is an integer decimal exponent. Any number should contain up to 12
characters. If the sign $s$ is omitted, the plus sign is assumed. The
exponent part is optional. If a number contains spaces, the spaces are
ignored.

If a card has the asterisk `\verb|*|' in the column 1, this card is
considered as a comment and ignored. Besides, if the first character in
the field 3 or 5 is the dollar sign `\verb|$|', all characters from the
dollar sign to the end of card are considered as a comment and ignored.

\pagebreak

MPS file should contain cards in the following order:

$\bullet$ NAME indicator card;

$\bullet$ ROWS indicator card;

$\bullet$ data cards specifying rows (constraints);

$\bullet$ COLUMNS indicator card;

$\bullet$ data cards specifying columns (structural variables) and
constraint coefficients;

$\bullet$ RHS indicator card;

$\bullet$ data cards specifying right-hand sides of constraints;

$\bullet$ RANGES indicator card;

$\bullet$ data cards specifying ranges for double-bounded constraints;

$\bullet$ BOUNDS indicator card;

$\bullet$ data cards specifying types and bounds of structural
variables;

$\bullet$ ENDATA indicator card.

{\it Section} is a group of cards consisting of an indicator card and
data cards succeeding this indicator card. For example, the ROWS section
consists of the ROWS indicator card and data cards specifying rows.

The sections RHS, RANGES, and BOUNDS are optional and may be omitted.

\section{Free MPS Format}

{\it Free MPS format} is an improved version of the standard (fixed)
MPS format described above.\footnote{This format was developed in the
beginning of 1990's by IBM as an alternative to the standard fixed MPS
format for Optimization Subroutine Library (OSL).} Note that all
changes in free MPS format concern only the coding of data while the
structure of data is the same for both fixed and free versions of the
MPS format.

In free MPS format indicator and data records\footnote{{\it Record} in
free MPS format has the same meaning as {\it card} in fixed MPS format.}
may have arbitrary length not limited to 80 characters. Fields of data
records have no predefined positions, i.e. the fields may begin in any
position, except position 1, which must be blank, and must be separated
from each other by one or more blanks. However, the fields must appear
in the same order as in fixed MPS format.

Symbolic names in fields 2, 3, and 5 may be longer than 8
characters\footnote{GLPK allows symbolic names having up to 255
characters.}
and must not contain embedded blanks.

Numeric values in fields 4 and 6 are limited to 12 characters and must
not contain embedded blanks.

Only six fields on each data record are used. Any other fields are
ignored.

If the first character of any field (not necessarily fields 3 and 5)
is the dollar sign (\$), all characters from the dollar sign to the end
of record are considered as a comment and ignored.

\section{NAME indicator card}

The NAME indicator card should be the first card in the MPS file (except
optional comment cards, which may precede the NAME card). This card
should contain the word \verb|NAME| in the columns 1---4 and the problem
name in the field 3. The problem name is optional and may be omitted.

\section{ROWS section}
\label{secrows}

The ROWS section should start with the indicator card, which contains
the word \verb|ROWS| in the columns 1---4.

Each data card in the ROWS section specifies one row (constraint) of the
problem. All these data cards have the following format.

`\verb|N|' in the field 1 means that the row is free (unbounded):
$$-\infty < x_i = a_{i1}x_{m+1} + a_{i2}x_{m+2} + \dots + a_{in}x_{m+n}
< +\infty;$$

`\verb|L|' in the field 1 means that the row is of ``less than or equal
to'' type:
$$-\infty < x_i = a_{i1}x_{m+1} + a_{i2}x_{m+2} + \dots + a_{in}x_{m+n}
\leq b_i;$$

`\verb|G|' in the field 1 means that the row is of ``greater than or
equal to'' type:
$$b_i \leq x_i = a_{i1}x_{m+1} + a_{i2}x_{m+2} + \dots + a_{in}x_{m+n}
< +\infty;$$

`\verb|E|' in the field 1 means that the row is of ``equal to'' type:
$$x_i = a_{i1}x_{m+1} + a_{i2}x_{m+2} + \dots + a_{in}x_{m+n} \leq
b_i,$$
where $b_i$ is a right-hand side. Note that each constraint has a
corresponding implictly defined auxiliary variable ($x_i$ above), whose
value is a value of the corresponding linear form, therefore row bounds
can be considered as bounds of such auxiliary variable.

The filed 2 specifies a row name (which is considered as the name of
the corresponding auxiliary variable).

The fields 3, 4, 5, and 6 are not used and should be empty.

Numerical values of all non-zero right-hand sides $b_i$ should be
specified in the RHS section (see below). All double-bounded (ranged)
constraints should be specified in the RANGES section (see below).

\section{COLUMNS section}

The COLUMNS section should start with the indicator card, which contains
the word \verb|COLUMNS| in the columns 1---7.

Each data card in the COLUMNS section specifies one or two constraint
coefficients $a_{ij}$ and also introduces names of columns, i.e. names
of structural variables. All these data cards have the following format.

The field 1 is not used and should be empty.

The field 2 specifies a column name. If this field is empty, the column
name from the immediately preceeding data card is assumed.

The field 3 specifies a row name defined in the ROWS section.

The field 4 specifies a numerical value of the constraint coefficient
$a_{ij}$, which is placed in the corresponding row and column.

The fields 5 and 6 are optional. If they are used, they should contain
a second pair ``row name---constraint coefficient'' for the same column.

Elements of the constraint matrix (i.e. constraint coefficients) should
be enumerated in the column wise manner: all elements for the current
column should be specified before elements for the next column. However,
the order of rows in the COLUMNS section may differ from the order of
rows in the ROWS section.

Constraint coefficients not specified in the COLUMNS section are
considered as zeros. Therefore zero coefficients may be omitted,
although it is allowed to explicitly specify them.

\section{RHS section}

The RHS section should start with the indicator card, which contains the
word \verb|RHS| in the columns 1---3.

Each data card in the RHS section specifies one or two right-hand sides
$b_i$ (see Section \ref{secrows}, page \pageref{secrows}). All these
data cards have the following format.

The field 1 is not used and should be empty.

The field 2 specifies a name of the right-hand side (RHS)
vector\footnote{This feature allows the user to specify several RHS
vectors in the same MPS file. However, before solving the problem a
particular RHS vector should be chosen.}. If this field is empty, the
RHS vector name from the immediately preceeding data card is assumed.

The field 3 specifies a row name defined in the ROWS section.

The field 4 specifies a right-hand side $b_i$ for the row, whose name is
specified in the field 3. Depending on the row type $b_i$ is a lower
bound (for the row of \verb|G| type), an upper bound (for the row of
\verb|L| type), or a fixed value (for the row of \verb|E|
type).\footnote{If the row is of {\tt N} type, $b_i$ is considered as
a constant term of the corresponding linear form. Should note, however,
this convention is non-standard.}

The fields 5 and 6 are optional. If they are used, they should contain
a second pair ``row name---right-hand side'' for the same RHS vector.

All right-hand sides for the current RHS vector should be specified
before right-hand sides for the next RHS vector. However, the order of
rows in the RHS section may differ from the order of rows in the ROWS
section.

Right-hand sides not specified in the RHS section are considered as
zeros. Therefore zero right-hand sides may be omitted, although it is
allowed to explicitly specify them.

\section{RANGES section}

The RANGES section should start with the indicator card, which contains
the word \verb|RANGES| in the columns 1---6.

Each data card in the RANGES section specifies one or two ranges for
double-side constraints, i.e. for constraints that are of the types
\verb|L| and \verb|G| at the same time:
$$l_i \leq x_i = a_{i1}x_{m+1} + a_{i2}x_{m+2} + \dots + a_{in}x_{m+n}
\leq u_i,$$
where $l_i$ is a lower bound, $u_i$ is an upper bound. All these data
cards have the following format.

The field 1 is not used and should be empty.

The field 2 specifies a name of the range vector\footnote{This feature
allows the user to specify several range vectors in the same MPS file.
However, before solving the problem a particular range vector should be
chosen.}. If this field is empty, the range vector name from the
immediately preceeding data card is assumed.

The field 3 specifies a row name defined in the ROWS section.

The field 4 specifies a range value $r_i$ (see the table below) for the
row, whose name is specified in the field 3.

The fields 5 and 6 are optional. If they are used, they should contain
a second pair ``row name---range value'' for the same range vector.

All range values for the current range vector should be specified before
range values for the next range vector. However, the order of rows in
the RANGES section may differ from the order of rows in the ROWS
section.

For each double-side constraint specified in the RANGES section its
lower and upper bounds are determined as follows:

\begin{center}
\begin{tabular}{cccc}
Row type & Sign of $r_i$ & Lower bound & Upper bound \\
\hline
{\tt G} & $+$ or $-$ & $b_i$ & $b_i + |r_i|$ \\
{\tt L} & $+$ or $-$ & $b_i - |r_i|$ & $b_i$ \\
{\tt E} & $+$ & $b_i$ & $b_i + |r_i|$ \\
{\tt E} & $-$ & $b_i - |r_i|$ & $b_i$ \\
\end{tabular}
\end{center}

\noindent
where $b_i$ is a right-hand side specified in the RHS section (if $b_i$
is not specified, it is considered as zero), $r_i$ is a range value
specified in the RANGES section.

\section{BOUNDS section}
\label{secbounds}

The BOUNDS section should start with the indicator card, which contains
the word \verb|BOUNDS| in the columns 1---6.

Each data card in the BOUNDS section specifies one (lower or upper)
bound for one structural variable (column). All these data cards have
the following format.

The indicator in the field 1 specifies the bound type:

\begin{tabular}{@{}ll}
\verb|LO| & lower bound; \\
\verb|UP| & upper bound; \\
\verb|FX| & fixed variable (lower and upper bounds are equal); \\
\verb|FR| & free variable (no bounds); \\
\verb|MI| & no lower bound (lower bound is ``minus infinity''); \\
\verb|PL| & no upper bound (upper bound is ``plus infinity''); \\
\end{tabular}

The field 2 specifies a name of the bound vector\footnote{This feature
allows the user to specify several bound vectors in the same MPS file.
However, before solving the problem a particular bound vector should be
chosen.}. If this field is empty, the bound vector name from the
immediately preceeding data card is assumed.

The field 3 specifies a column name defined in the COLUMNS section.

The field 4 specifies a bound value. If the bound type in the field 1
differs from \verb|LO|, \verb|UP|, and \verb|FX|, the value in the field
4 is ignored and may be omitted.

The fields 5 and 6 are not used and should be empty.

All bound values for the current bound vector should be specified before
bound values for the next bound vector. However, the order of columns in
the BOUNDS section may differ from the order of columns in the COLUMNS
section. Specification of a lower bound should precede specification of
an upper bound for the same column (if both the lower and upper bounds
are explicitly specified).

By default, all columns (structural variables) are non-negative, i.e.
have zero lower bound and no upper bound. Lower ($l_j$) and upper
($u_j$) bounds of some column (structural variable $x_j$) are set in the
following way, where $s_j$ is a corresponding bound value explicitly
specified in the BOUNDS section:

\begin{tabular}{@{}ll}
\verb|LO| & sets $l_j$ to $s_j$; \\
\verb|UP| & sets $u_j$ to $s_j$; \\
\verb|FX| & sets both $l_j$ and $u_j$ to $s_j$; \\
\verb|FR| & sets $l_j$ to $-\infty$ and $u_j$ to $+\infty$; \\
\verb|MI| & sets $l_j$ to $-\infty$; \\
\verb|PL| & sets $u_j$ to $+\infty$. \\
\end{tabular}

\section{ENDATA indicator card}

The ENDATA indicator card should be the last card of MPS file (except
optional comment cards, which may follow the ENDATA card). This card
should contain the word \verb|ENDATA| in the columns 1---6.

\section{Specifying objective function}

It is impossible to explicitly specify the objective function and
optimization direction in the MPS file. However, the following implicit
rule is used by default: the first row of \verb|N| type is considered
as a row of the objective function (i.e. the objective function is the
corresponding auxiliary variable), which should be {\it minimized}.

GLPK also allows specifying a constant term of the objective function
as a right-hand side of the corresponding row in the RHS section.

\section{Example of MPS file}
\label{secmpsex}

In order to illustrate what the MPS format is, consider the following
example of LP problem:

\medskip
\noindent minimize
$$
value = .03\ bin_1 + .08\ bin_2 + .17\ bin_3 + .12\ bin_4 + .15\ bin_5
+ .21\ alum + .38\ silicon
$$

\noindent subject to linear constraints
$$
\begin{array}{ll}
yield &= \ \ \ \ \;bin_1 + \ \ \ \ \;bin_2 + \ \ \ \ \;bin_3 +
         \ \ \ \ \;bin_4 + \ \ \ \ \;bin_5 + \ \ \ \ \;alum +
         \ \ \ \ \;silicon \\
fe    &= .15\ bin_1 + .04\ bin_2 + .02\ bin_3 + .04\ bin_4 + .02\ bin_5
         + .01\ alum + .03\ silicon \\
cu    &= .03\ bin_1 + .05\ bin_2 + .08\ bin_3 + .02\ bin_4 + .06\ bin_5
         + .01\ alum \\
mn    &= .02\ bin_1 + .04\ bin_2 + .01\ bin_3 + .02\ bin_4 + .02\ bin_5
         \\
mg    &= .02\ bin_1 + .03\ bin_2
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ + .01\ bin_5 \\
al    &= .70\ bin_1 + .75\ bin_2 + .80\ bin_3 + .75\ bin_4 + .80\ bin_5
         + .97\ alum \\
si    &= .02\ bin_1 + .06\ bin_2 + .08\ bin_3 + .12\ bin_4 + .02\ bin_5
         + .01\ alum + .97\ silicon \\
\end{array}
$$

\noindent and bounds of (auxiliary and structural) variables
$$
\begin{array}{r@{\ }l@{\ }l@{\ }l@{\ }rcr@{\ }l@{\ }l@{\ }l@{\ }r}
&&yield&=&2000&&0&\leq&bin_1&\leq&200\\
-\infty&<&fe&\leq&60&&0&\leq&bin_2&\leq&2500\\
-\infty&<&cu&\leq&100&&400&\leq&bin_3&\leq&800\\
-\infty&<&mn&\leq&40&&100&\leq&bin_4&\leq&700\\
-\infty&<&mg&\leq&30&&0&\leq&bin_5&\leq&1500\\
1500&\leq&al&<&+\infty&&0&\leq&alum&<&+\infty\\
250&\leq&si&\leq&300&&0&\leq&silicon&<&+\infty\\
\end{array}
$$

A complete MPS file which specifies data for this example is shown
below (the first two comment lines show card positions).

\begin{verbatim}
*000000001111111111222222222233333333334444444444555555555566
*234567890123456789012345678901234567890123456789012345678901
NAME          PLAN
ROWS
 N  VALUE
 E  YIELD
 L  FE
 L  CU
 L  MN
 L  MG
 G  AL
 L  SI
COLUMNS
    BIN1      VALUE           .03000   YIELD          1.00000
              FE              .15000   CU              .03000
              MN              .02000   MG              .02000
              AL              .70000   SI              .02000
    BIN2      VALUE           .08000   YIELD          1.00000
              FE              .04000   CU              .05000
              MN              .04000   MG              .03000
              AL              .75000   SI              .06000
    BIN3      VALUE           .17000   YIELD          1.00000
              FE              .02000   CU              .08000
              MN              .01000   AL              .80000
              SI              .08000
    BIN4      VALUE           .12000   YIELD          1.00000
              FE              .04000   CU              .02000
              MN              .02000   AL              .75000
              SI              .12000
    BIN5      VALUE           .15000   YIELD          1.00000
              FE              .02000   CU              .06000
              MN              .02000   MG              .01000
              AL              .80000   SI              .02000
    ALUM      VALUE           .21000   YIELD          1.00000
              FE              .01000   CU              .01000
              AL              .97000   SI              .01000
    SILICON   VALUE           .38000   YIELD          1.00000
              FE              .03000   SI              .97000
RHS
    RHS1      YIELD       2000.00000   FE            60.00000
              CU           100.00000   MN            40.00000
              SI           300.00000
              MG            30.00000   AL          1500.00000
RANGES
    RNG1      SI            50.00000
BOUNDS
 UP BND1      BIN1         200.00000
 UP           BIN2        2500.00000
 LO           BIN3         400.00000
 UP           BIN3         800.00000
 LO           BIN4         100.00000
 UP           BIN4         700.00000
 UP           BIN5        1500.00000
ENDATA
\end{verbatim}

\section{MIP features}

The MPS format provides two ways for introducing integer variables into
the problem.

The first way is most general and based on using special marker cards
INTORG and INTEND. These marker cards are placed in the COLUMNS section.
The INTORG card indicates the start of a group of integer variables
(columns), and the card INTEND indicates the end of the group. The MPS
file may contain arbitrary number of the marker cards.

The marker cards have the same format as the data cards (see Section
\ref{secmps}, page \pageref{secmps}).

The fields 1, 2, and 6 are not used and should be empty.

The field 2 should contain a marker name. This name may be arbitrary.

The field 3 should contain the word \verb|'MARKER'| (including
apostrophes).

The field 5 should contain either the word \verb|'INTORG'| (including
apostrophes) for the marker card, which begins a group of integer
columns, or the word \verb|'INTEND'| (including apostrophes) for the
marker card, which ends the group.

The second way is less general but more convenient in some cases. It
allows the user to declare integer columns using two additional types
of bounds, which are specified in the field 1 of data cards in the
BOUNDS section (see Section \ref{secbounds}, page \pageref{secbounds}):

\begin{tabular}{@{}lp{135mm}@{}}
\verb|UI| & upper integer. This bound type specifies that the
corresponding column (structural variable), whose name is specified in
the field 3, is of integer kind. In this case an upper bound of the
column should be specified in the field 4 (like in the case of \verb|UP|
bound type). \\
\verb|BV| & binary variable. This bound type specifies that the
corresponding column (structural variable), whose name is specified in
the field 3, is of integer kind, its lower bound is zero, and its upper
bound is one (thus, such variable being of integer kind can have only
two values zero and one). In this case a numeric value specified in the
field 4 is ignored and may be omitted.
\end{tabular}

Consider the following example of MIP problem:

\medskip

\noindent
\hspace{1in} minimize
$$Z = 3 x_1 + 7 x_2 - x_3 + x4$$
\hspace{1in} subject to linear constraints
$$
\begin{array}{c}
\nonumber r_1 = 2   x_1 - \ \ x_2 + \ \ x_3 - \ \;x_4 \\
\nonumber r_2 = \ \;x_1 - \ \;x_2 - 6   x_3 + 4   x_4 \\
\nonumber r_3 = 5   x_1 +   3 x_2 \ \ \ \ \ \ \ \ \ + \ \ x_4 \\
\end{array}
$$
\hspace{1in} and bound of variables
$$
\begin{array}{cccl}
\nonumber 1 \leq r_1 < +\infty && 0 \leq x_1 \leq 4 &{\rm(continuous)}\\
\nonumber 8 \leq r_2 < +\infty && 2 \leq x_2 \leq 5 &{\rm(integer)}   \\
\nonumber 5 \leq r_3 < +\infty && 0 \leq x_3 \leq 1 &{\rm(integer)}   \\
\nonumber                      && 3 \leq x_4 \leq 8 &{\rm(continuous)}\\
\end{array}
$$

The corresponding MPS file may look like the following:

\begin{verbatim}
NAME          SAMP1
ROWS
 N  Z
 G  R1
 G  R2
 G  R3
COLUMNS
    X1        R1                2.0    R2                 1.0
    X1        R3                5.0    Z                  3.0
    MARK0001  'MARKER'                 'INTORG'
    X2        R1               -1.0    R2                -1.0
    X2        R3                3.0    Z                  7.0
    X3        R1                1.0    R2                -6.0
    X3        Z                -1.0
    MARK0002  'MARKER'                 'INTEND'
    X4        R1               -1.0    R2                 4.0
    X4        R3                1.0    Z                  1.0
RHS
    RHS1      R1                1.0
    RHS1      R2                8.0
    RHS1      R3                5.0
BOUNDS
 UP BND1      X1                4.0
 LO BND1      X2                2.0
 UP BND1      X2                5.0
 UP BND1      X3                1.0
 LO BND1      X4                3.0
 UP BND1      X4                8.0
ENDATA
\end{verbatim}

The same example may be coded without INTORG/INTEND markers using the
bound type UI for the variable $x_2$ and the bound type BV for the
variable $x_3$:

\begin{verbatim}
NAME          SAMP2
ROWS
 N  Z
 G  R1
 G  R2
 G  R3
COLUMNS
    X1        R1                2.0    R2                 1.0
    X1        R3                5.0    Z                  3.0
    X2        R1               -1.0    R2                -1.0
    X2        R3                3.0    Z                  7.0
    X3        R1                1.0    R2                -6.0
    X3        Z                -1.0
    X4        R1               -1.0    R2                 4.0
    X4        R3                1.0    Z                  1.0
RHS
    RHS1      R1                1.0
    RHS1      R2                8.0
    RHS1      R3                5.0
BOUNDS
 UP BND1      X1                4.0
 LO BND1      X2                2.0
 UI BND1      X2                5.0
 BV BND1      X3
 LO BND1      X4                3.0
 UP BND1      X4                8.0
ENDATA
\end{verbatim}

\section{Specifying predefined basis}
\label{secbas}

The MPS format can also be used to specify some predefined basis for an
LP problem, i.e. to specify which rows and columns are basic and which
are non-basic.

The order of a basis file in the MPS format is:

$\bullet$ NAME indicator card;

$\bullet$ data cards (can appear in arbitrary order);

$\bullet$ ENDATA indicator card.

Each data card specifies either a pair "basic column---non-basic row"
or a non-basic column. All the data cards have the following format.

`\verb|XL|' in the field 1 means that a column, whose name is given in
the field 2, is basic, and a row, whose name is given in the field 3,
is non-basic and placed on its lower bound.

`\verb|XU|' in the field 1 means that a column, whose name is given in
the field 2, is basic, and a row, whose name is given in the field 3,
is non-basic and placed on its upper bound.

`\verb|LL|' in the field 1 means that a column, whose name is given in
the field 3, is non-basic and placed on its lower bound.

`\verb|UL|' in the field 1 means that a column, whose name is given in
the field 3, is non-basic and placed on its upper bound.

The field 2 contains a column name.

If the indicator given in the field 1 is `\verb|XL|' or `\verb|XU|',
the field 3 contains a row name. Otherwise, if the indicator is
`\verb|LL|' or `\verb|UL|', the field 3 is not used and should be empty.

The field 4, 5, and 6 are not used and should be empty.

A basis file in the MPS format acts like a patch: it doesn't specify
a basis completely, instead that it is just shows in what a given basis
differs from the "standard" basis, where all rows (auxiliary variables)
are assumed to be basic and all columns (structural variables) are
assumed to be non-basic.

As an example here is a basis file that specifies an optimal basis
for the example LP problem given in Section \ref{secmpsex},
Page \pageref{secmpsex}:

\begin{verbatim}
*000000001111111111222222222233333333334444444444555555555566
*234567890123456789012345678901234567890123456789012345678901
NAME          PLAN
 XL BIN2      YIELD
 XL BIN3      FE
 XL BIN4      MN
 XL ALUM      AL
 XL SILICON   SI
 LL BIN1
 LL BIN5
ENDATA
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{CPLEX LP Format}
\label{chacplex}

\section{Prelude}

The CPLEX LP format\footnote{The CPLEX LP format was developed in
the end of 1980's by CPLEX Optimization, Inc. as an input format for
the CPLEX linear programming system. Although the CPLEX LP format is
not as widely used as the MPS format, being row-oriented it is more
convenient for coding mathematical programming models by human. This
appendix describes only the features of the CPLEX LP format which are
implemented in the GLPK package.} is intended for coding LP/MIP problem
data. It is a row-oriented format that assumes the formulation of LP/MIP
problem (1.1)---(1.3) (see Section \ref{seclp}, page \pageref{seclp}).

CPLEX LP file is a plain text file written in CPLEX LP format. Each text
line of this file may contain up to 255 characters\footnote{GLPK allows
text lines of arbitrary length.}. Blank lines are ignored. If a line
contains the backslash character ($\backslash$), this character and
everything that follows it until the end of line are considered as a
comment and also ignored.

An LP file is coded by the user using the following elements:

$\bullet$ keywords;

$\bullet$ symbolic names;

$\bullet$ numeric constants;

$\bullet$ delimiters;

$\bullet$ blanks.

{\it Keywords} that may be used in the LP file are the following:

\begin{verbatim}
      minimize        minimum        min
      maximize        maximum        max
      subject to      such that      s.t.      st.      st
      bounds          bound
      general         generals       gen
      integer         integers       int
      binary          binaries       bin
      infinity        inf
      free
      end
\end{verbatim}

\noindent
All the keywords are case insensitive. Keywords given above on the same
line are equivalent. Any keyword (except \verb|infinity|, \verb|inf|,
and \verb|free|) being used in the LP file must start at the beginning
of a text line.

{\it Symbolic names} are used to identify the objective function,
constraints (rows), and variables (columns). All symbolic names are case
sensitive and may contain up to 16 alphanumeric characters\footnote{GLPK
allows symbolic names having up to 255 characters.} (\verb|a|, \dots,
\verb|z|, \verb|A|, \dots, \verb|Z|, \verb|0|, \dots, \verb|9|) as well
as the following characters:

\begin{verbatim}
      !  "  #  $  %  &  (  )  /  ,  .  ;  ?  @  _  `  '  {  }  |  ~
\end{verbatim}

\noindent
with exception that no symbolic name can begin with a digit or a period.

{\it Numeric constants} are used to denote constraint and objective
coefficients, right-hand sides of constraints, and bounds of variables.
They are coded in the standard form $xx$\verb|E|$syy$, where $xx$ is
a real number with optional decimal point, $s$ is a sign (\verb|+| or
\verb|-|), $yy$ is an integer decimal exponent. Numeric constants may
contain arbitrary number of characters. The exponent part is optional.
The letter `\verb|E|' can be coded as `\verb|e|'. If the sign $s$ is
omitted, plus is assumed.

{\it Delimiters} that may be used in the LP file are the following:

\begin{verbatim}
      :
      +
      -
      <   <=   =<
      >   >=   =>
      =
\end{verbatim}

\noindent
Delimiters given above on the same line are equivalent. The meaning of
the delimiters will be explained below.

{\it Blanks} are non-significant characters. They may be used freely to
improve readability of the LP file. Besides, blanks should be used to
separate elements from each other if there is no other way to do that
(for example, to separate a keyword from a following symbolic name).

The order of an LP file is:

$\bullet$ objective function definition;

$\bullet$ constraints section;

$\bullet$ bounds section;

$\bullet$ general, integer, and binary sections (can appear in arbitrary
order);

$\bullet$ end keyword.

These components are discussed in following sections.

\section{Objective function definition}

The objective function definition must appear first in the LP file. It
defines the objective function and specifies the optimization direction.

The objective function definition has the following form:
$$
\left\{
\begin{array}{@{}c@{}}
{\tt minimize} \\ {\tt maximize}
\end{array}
\right\}\ f\ {\tt :}\ s\ c\ x\ s\ c\ x\ \dots\ s\ c\ x
$$
where $f$ is a symbolic name of the objective function, $s$ is a sign
\verb|+| or \verb|-|, $c$ is a numeric constant that denotes an
objective coefficient, $x$ is a symbolic name of a variable.

If necessary, the objective function definition can be continued on as
many text lines as desired.

The name of the objective function is optional and may be omitted
(together with the semicolon that follows it). In this case the default
name `\verb|obj|' is assigned to the objective function.

If the very first sign $s$ is omitted, the sign plus is assumed. Other
signs cannot be omitted.

If some objective coefficient $c$ is omitted, 1 is assumed.

Symbolic names $x$ used to denote variables are recognized by context
and therefore needn't to be declared somewhere else.

Here is an example of the objective function definition:

\begin{verbatim}
      Minimize Z : - x1 + 2 x2 - 3.5 x3 + 4.997e3x(4) + x5 + x6 +
         x7 - .01x8
\end{verbatim}

\section{Constraints section}

The constraints section must follow the objective function definition.
It defines a system of equality and/or inequality constraints.

The constraint section has the following form:

\begin{center}
\begin{tabular}{l}
\verb|subject to| \\
{\it constraint}$_1$ \\
{\it constraint}$_2$ \\
\hspace{20pt}\dots \\
{\it constraint}$_m$ \\
\end{tabular}
\end{center}

\noindent where {\it constraint}$_i, i=1,\dots,m,$ is a particular
constraint definition.

Each constraint definition can be continued on as many text lines as
desired. However, each constraint definition must begin on a new line
except the very first constraint definition which can begin on the same
line as the keyword `\verb|subject to|'.

Constraint definitions have the following form:
$$
r\ {\tt:}\ s\ c\ x\ s\ c\ x\ \dots\ s\ c\ x
\ \left\{
\begin{array}{@{}c@{}}
\mbox{\tt<=} \\ \mbox{\tt>=} \\ \mbox{\tt=}
\end{array}
\right\}\ b
$$
where $r$ is a symbolic name of a constraint, $s$ is a sign \verb|+| or
\verb|-|, $c$ is a numeric constant that denotes a constraint
coefficient, $x$ is a symbolic name of a variable, $b$ is a right-hand
side.

The name $r$ of a constraint (which is the name of the corresponding
auxiliary variable) is optional and may be omitted (together with the
semicolon that follows it). In this case the default names like
`\verb|r.nnn|' are assigned to unnamed constraints.

The linear form $s\ c\ x\ s\ c\ x\ \dots\ s\ c\ x$ in the left-hand
side of a constraint definition has exactly the same meaning as in the
case of the objective function definition (see above).

After the linear form one of the following delimiters that indicate
the constraint sense must be specified:

\verb|<=| \ means `less than or equal to'

\verb|>=| \ means `greater than or equal to'

\verb|= | \ means `equal to'

The right hand side $b$ is a numeric constant with an optional sign.

Here is an example of the constraints section:

\begin{verbatim}
      Subject To
         one: y1 + 3 a1 - a2 - b >= 1.5
         y2 + 2 a3 + 2
            a4 - b >= -1.5
         two : y4 + 3 a1 + 4 a5 - b <= +1
         .20y5 + 5 a2 - b = 0
         1.7 y6 - a6 + 5 a777 - b >= 1
\end{verbatim}

(Should note that it is impossible to express ranged constraints in the
CPLEX LP format. Each a ranged constraint can be coded as two
constraints with identical linear forms in the left-hand side, one of
which specifies a lower bound and other does an upper one of the
original ranged constraint.)

\section{Bounds section}

The bounds section is intended to define bounds of variables. This
section is optional; if it is specified, it must follow the constraints
section. If the bound section is omitted, all variables are assumed to
be non-negative (i.e. that they have zero lower bound and no upper
bound).

The bounds section has the following form:

\begin{center}
\begin{tabular}{l}
\verb|bounds| \\
{\it definition}$_1$ \\
{\it definition}$_2$ \\
\hspace{20pt}\dots \\
{\it definition}$_p$ \\
\end{tabular}
\end{center}

\noindent
where {\it definition}$_k, k=1,\dots,p,$ is a particular bound
definition.

Each bound definition must begin on a new line\footnote{The GLPK
implementation allows several bound definitions to be placed on the same
line.} except the very first bound definition which can begin on the
same line as the keyword `\verb|bounds|'.

Syntactically constraint definitions can have one of the following six
forms:

\begin{center}
\begin{tabular}{ll}
$x$ \verb|>=| $l$ &              specifies a lower bound \\
$l$ \verb|<=| $x$ &              specifies a lower bound \\
$x$ \verb|<=| $u$ &              specifies an upper bound \\
$l$ \verb|<=| $x$ \verb|<=| $u$ &specifies both lower and upper bounds\\
$x$ \verb|=| $t$                &specifies a fixed value \\
$x$ \verb|free|                 &specifies free variable
\end{tabular}
\end{center}

\noindent
where $x$ is a symbolic name of a variable, $l$ is a numeric constant
with an optional sign that defines a lower bound of the variable or
\verb|-inf| that means that the variable has no lower bound, $u$ is a
numeric constant with an optional sign that defines an upper bound of
the variable or \verb|+inf| that means that the variable has no upper
bound, $t$ is a numeric constant with an optional sign that defines a
fixed value of the variable.

\pagebreak

By default all variables are non-negative, i.e. have zero lower bound
and no upper bound. Therefore definitions of these default bounds can be
omitted in the bounds section.

Here is an example of the bounds section:

\begin{verbatim}
      Bounds
         -inf <= a1 <= 100
         -100 <= a2
         b <= 100
         x2 = +123.456
         x3 free
\end{verbatim}

\section{General, integer, and binary sections}

The general, integer, and binary sections are intended to define
some variables as integer or binary. All these sections are optional
and needed only in case of MIP problems. If they are specified, they
must follow the bounds section or, if the latter is omitted, the
constraints section.

All the general, integer, and binary sections have the same form as
follows:

\begin{center}
\begin{tabular}{l}
$
\left\{
\begin{array}{@{}l@{}}
\verb|general| \\
\verb|integer| \\
\verb|binary | \\
\end{array}
\right\}
$ \\
\hspace{10pt}$x_1$ \\
\hspace{10pt}$x_2$ \\
\hspace{10pt}\dots \\
\hspace{10pt}$x_q$ \\
\end{tabular}
\end{center}

\noindent
where $x_k$ is a symbolic name of variable, $k=1,\dots,q$.

Each symbolic name must begin on a new line\footnote{The GLPK
implementation allows several symbolic names to be placed on the same
line.} except the very first symbolic name which can begin on the same
line as the keyword `\verb|general|', `\verb|integer|', or
`\verb|binary|'.

If a variable appears in the general or the integer section, it is
assumed to be general integer variable. If a variable appears in the
binary section, it is assumed to be binary variable, i.e. an integer
variable whose lower bound is zero and upper bound is one. (Note that
if bounds of a variable are specified in the bounds section and then
the variable appears in the binary section, its previously specified
bounds are ignored.)

Here is an example of the integer section:

\begin{verbatim}
      Integer
         z12
         z22
         z35
\end{verbatim}

\section{End keyword}

The keyword `\verb|end|' is intended to end the LP file. It must begin
on a separate line and no other elements (except comments and blank
lines) must follow it. Although this keyword is optional, it is strongly
recommended to include it in the LP file.

\section{Example of CPLEX LP file}

Here is a complete example of CPLEX LP file that corresponds to the
example given in Section \ref{secmpsex}, page \pageref{secmpsex}.

\begin{verbatim}
\* plan.lp *\

Minimize
   value: .03 bin1 + .08 bin2 + .17 bin3 + .12 bin4 + .15 bin5 +
          .21 alum + .38 silicon

Subject To
   yield:     bin1 +     bin2 +     bin3 +     bin4 +     bin5 +
              alum +     silicon                                 =  2000

   fe:    .15 bin1 + .04 bin2 + .02 bin3 + .04 bin4 + .02 bin5 +
          .01 alum + .03 silicon                                 <=   60

   cu:    .03 bin1 + .05 bin2 + .08 bin3 + .02 bin4 + .06 bin5 +
          .01 alum                                               <=  100

   mn:    .02 bin1 + .04 bin2 + .01 bin3 + .02 bin4 + .02 bin5   <=   40

   mg:    .02 bin1 + .03 bin2                       + .01 bin5   <=   30

   al:    .70 bin1 + .75 bin2 + .80 bin3 + .75 bin4 + .80 bin5 +
          .97 alum                                               >= 1500

   si1:   .02 bin1 + .06 bin2 + .08 bin3 + .12 bin4 + .02 bin5 +
          .01 alum + .97 silicon                                 >=  250

   si2:   .02 bin1 + .06 bin2 + .08 bin3 + .12 bin4 + .02 bin5 +
          .01 alum + .97 silicon                                 <=  300

Bounds
          bin1 <=  200
          bin2 <= 2500
   400 <= bin3 <=  800
   100 <= bin4 <=  700
          bin5 <= 1500

End

\* eof *\
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Stand-alone LP/MIP Solver}
\label{chaglpsol}

The GLPK package includes the program \verb|glpsol| which is a
stand-alone LP/MIP solver. This program can be invoked from the command
line of from the shell to read LP/MIP problem data in any format
supported by GLPK, solve the problem, and write the obtained problem
solution to a text file in plain format.

\bigskip

\noindent
{\bf Usage}

\bigskip

\noindent
\verb|glpsol| [{\it options\dots}] [{\it filename}]

\bigskip

\noindent
{\bf General options}

\bigskip

\noindent
\begin{tabular}{@{}p{30mm}p{115.8mm}@{}}
\verb|--glp|      &  read LP/MIP model in GNU LP format \\
\verb|--mps|      &  read LP/MIP problem in fixed MPS format (default)\\
\verb|--freemps|  &  read LP/MIP problem in free MPS format \\
\verb|--cpxlp|    &  read LP/MIP problem in CPLEX LP format \\
\verb|--math|     &  read LP/MIP model written in GNU MathProg modeling
                     language \\
\multicolumn{2}{@{}l}{{\tt -m} {\it filename}, {\tt --model}
{\it filename}} \\
                  &  read model section and optional data section from
                     filename (the same as \verb|--math|) \\
\multicolumn{2}{@{}l}{{\tt -d} {\it filename}, {\tt --data}
{\it filename}} \\
                  &  read data section from filename (for \verb|--math|
                     only); if model file also has data section, that
                     section is ignored \\
\multicolumn{2}{@{}l}{{\tt -y} {\it filename}, {\tt --display}
{\it filename}} \\
                  &  send display output to filename (for \verb|--math|
                     only); by default the output is sent to
                     \verb|stdout| \\
\verb|--min|      &  minimization \\
\verb|--max|      &  maximization \\
\verb|--scale|    &  scale problem (default) \\
\verb|--noscale|  &  do not scale problem \\
\verb|--simplex|  &  use simplex method (default) \\
\verb|--interior| &  use interior point method (for pure LP only) \\
\multicolumn{2}{@{}l}{{\tt -o} {\it filename}, {\tt --output}
{\it filename}} \\
                  &  write solution to filename in plain text format \\
\multicolumn{2}{@{}l}{{\tt --bounds} {\it filename}} \\
                  &  write sensitivity bounds to filename in plain text
                     format (LP only) \\
\end{tabular}

\noindent
\begin{tabular}{@{}p{30mm}p{115.8mm}@{}}
\verb|--tmlim| {\it nnn}
                  &  limit solution time to {\it nnn} seconds
                     (\verb|--tmlim 0| allows obtaining solution at
                     initial point) \\
\verb|--check|    &  do not solve problem, check input data only \\
\verb|--name| {\it probname}
                  &  change problem name to {\it probname} \\
\verb|--plain|    &  use plain names of rows and columns (default) \\
\verb|--orig|     &  try using original names of rows and columns
                     (default for \verb|--mps|) \\
\verb|--wglp| {\it filename}
                  &  write problem to {\it filename} in GNU LP format \\
\verb|--wmps| {\it filename}
                  &  write problem to {\it filename} in fixed MPS
                     format \\
\multicolumn{2}{@{}l}{{\tt --wfreemps} {\it filename}} \\
                  &  write problem to {\it filename} in free MPS
                     format \\
\multicolumn{2}{@{}l}{{\tt --wcpxlp} {\it filename}} \\
                  &  write problem to {\it filename} in CPLEX LP
                     format \\
\verb|--wtxt| {\it filename}
                  &  write problem to {\it filename} in plain text
                     format \\
\verb|-h|, \verb|--help|
                  &  display this help information and exit \\
\verb|-v|, \verb|--version|
                  &  display program version and exit \\
\end{tabular}

\bigskip

\noindent
{\bf Options specific to simplex method}

\bigskip

\noindent
\begin{tabular}{@{}p{30mm}p{115.8mm}@{}}
\verb|--std|      &  use standard initial basis of all slacks \\
\verb|--adv|      &  use advanced initial basis (default) \\
\verb|--bas| {\it filename}
                  &  read initial basis from {\it filename} in MPS
                     format \\
\verb|--steep|    &  use steepest edge technique (default) \\
\verb|--nosteep|  &  use standard ``textbook'' pricing \\
\verb|--relax|    &  use Harris' two-pass ratio test (default) \\
\verb|--norelax|  &  use standard ``textbook'' ratio test \\
\verb|--presol|   &  use LP presolver (default; assumes \verb|--scale|
                     and \verb|--adv|) \\
\verb|--nopresol| &  do not use LP presolver \\
\verb|--wbas| {\it filename}
                  & write final basis to {\it filename} in MPS format \\
\end{tabular}

\bigskip

\noindent
{\bf Options specific to MIP}

\bigskip

\noindent
\begin{tabular}{@{}p{30mm}p{115.8mm}@{}}
\verb|--nomip|    &  consider all integer variables as continuous
                     (allows solving MIP as pure LP) \\
\verb|--first|    &  branch on first integer variable \\
\verb|--last|     &  branch on last integer variable \\
\verb|--drtom|    &  branch using heuristic by Driebeck and Tomlin
                     (default) \\
\verb|--mostf|    &  branch on most fractional varaible \\
\verb|--dfs|      &  backtrack using depth first search \\
\verb|--bfs|      &  backtrack using breadth first search \\
\verb|--bestp|    &  backtrack using the best projection heuristic
                     (default) \\
\verb|--bestb|    &  backtrack using node with best local bound \\
\end{tabular}

\bigskip

\noindent
For description of the MPS format see Appendix \ref{champs},
page \pageref{champs}.

\bigskip

\noindent
For description of the CPLEX LP format see Appendix \ref{chacplex},
page \pageref{chacplex}.

\bigskip

\noindent
For description of the modeling language see the document ``GLPK:
Modeling Language GNU MathProg'' included in the GLPK distribution.

\end{document}

%% eof %%
