\documentclass[12pt,twoside,twocolumn]{article}
\usepackage{linuxexpo}
\begin{document}

\title{The Gnumeric spreadsheet: a test-bed for component programming.}

\author{
	\textit{Miguel de Icaza}\\
	Instituto de Ciencias Nucleares \\
	Universidad Nacional Aut\'onoma de M\'exico.\\
	\small
	miguel@gnu.org\\
}

\maketitle

\begin{abstract}
The Gnumeric spreadsheet is part of the GNOME project.  Gnumeric is
licensed under the terms of the GNU GPL.  I show how the main
components of Gnumeric work and how the Gnumeric engine is exported
trough CORBA and scripting language facilities to customize and extend
it.  A view at the development of Gnumeric and some lessons
learned are showed.
\end{abstract}



\section{Motivation.}

The Gnumeric spreadsheet is part of the GNU Network Object Model
Environment \cite{gnome} project.  The GNOME project provides a framework
for application development and components reuse.  The GNOME desktop
is a free[gnufree] user-friendly, consistent, easy to use yet powerful set
of applications and desktop integration tools for UNIX systems.

The lack of a standard, and freely available setup for component
programming is a problem present on the UNIX world nowadays.  One of
the issues being addressed by the GNOME project is providing such a
framework.  The GNOME framework is based on the CORBA[omg] object
model.  During the design of the GNOME component interfaces, we tried
to address a range of needs:

\begin{itemize}

	\item {Automation:} Allow applications to be remotely
	controlled.  People should be able to launch and control GNOME
	applications remotely.  This is achieved by the Gnome Object
	Activation Directory (GOAD), the supporting libraries (the
	GNORBA libraries) and making use of the CORBA facilities for
	binding CORBA to scripting languages.

	\item {Compound document creation:} It is important to design
	and implement the GNOME applications in such a way that they
	will let the user create compound documents:  Documents whose
	contents might have been produced by different tools. 

	\item {In-place document editing:} The next step to compound
	document creation is providing ways to edit embedded documents
	inside a container application.  This means that it should be
	possible to make changes to an embedded spreadsheet document
	inside a word processing document in a seamless fashion:  it
	is important to make this integration simple and easy to use. 

	\item {Component reuse:} Filters and pipes proved to be 
	important building blocks in Unix but they are very limited:
	The flow of control usually goes in a single direction and the
	protocol used to exchange information is very simple for
	today's needs.  

	\item {Desktop integration:} The GNOME desktop deals with
	CORBA interfaces to services:  This means that as far as an
	application is concerned, they only care about the published
	interface that a program provides.

	In GNOME, interfaces for specific tasks play an important
	role: it is up to the user to choose which implementation of
	those interfaces he uses (for example, the Mailer interface is
	implemented by the GNOME Balsa mail reader, but it can be
	implemented by Emacs RMAIL, Emacs GNUS or the Mozilla Mail
	Reader:  if any of those provide the CORBA Mailer interface
	they will fit and integrate in the GNOME desktop). 

\end{itemize}

Defining sensible interfaces for the above listed tasks without having
real applications that would use them is an experiment in futility.
One of the motivations for coding the Gnumeric spreadsheet were to
address the issues listed above.

The GNOME project would not be complete without a collection of
commercial-grade end-user applications, and Gnumeric is one of these
applications.  Gnumeric was designed and coded to be of
commercial-grade quality: the design of the code allows for Gnumeric
to be extended to what the commercial offering has these days.

One of Gnumeric goals was thus to become a replacement for the
commercial offerings and possibly go beyond their offerings.  Thus,
various design decisions in Gnumeric are shaped after the industry top
spreadsheets.  For example, the formats used to configure the way
numbers are displayed in Gnumeric mimics the Excel format.




\section{The display engine.}

The Gnumeric display engine is built on top of the GNOME
Canvas\cite{canvas}.  The GNOME canvas is an extensible display and event
dispatch engine that provides high-quality, flicker-free display
primitives to applications.  A GNOME canvas is a container for objects
of type GnomeCanvasItem which are the actual objects that receive
events and display themselves inside the canvas.

The GNOME canvas provides with a number of graphic primitives that can
be used to construct powerful applications.  All of these primitives
are derived from the GTK+ object GnomeCanvasItem.

The flicker-free display is achieved by having the GnomeCanvasItem
objects render their contents in an off-screen pixmap which is later
transfered in a single operation to the actual exposed region.  The
pixmap transfer takes place by default on the idle loop handler.  Even
when it is possible to force a redraw at any point, applications that
are user-input driven do not benefit from this feature.

The availability of a flick-free display engine frees the programmer
from having to deal with intricate paint setups for avoiding flicker
on the screen and concentrate on his code features.  Complex graphical
user interface are thus easy to construct with little hassle.  Display
Postscript based systems provide this same power to the user.

In the practice, we have noticed that Gnumeric flickers a lot less
than the commercial spreadsheets due to this engine in various cases
where a lot of changes are being done to the Gnumeric grid display. 

\subsection{Subclassing the Canvas objects.}

Gnumeric was the first GNOME application to use the GNOME Canvas
engine: it was also the first application to effectively provide its
own Canvas items to simplify the programming of all the graphical
elements that people expect from a modern spreadsheet.

The GNOME canvas provides to graphic application authors a powerful
abstraction for creating elaborate graphical applications.  All of the
details of event dispatching, rendering, flicker free display and
redrawing are done by the Canvas on behalf of the programmer.

The code in Gnumeric subclasses the GnomeCanvas widget engine to
provide spreadsheet specific event dispatching and to implement the
special key bindings and focus policies as expected on a spreadsheet.

The spreadsheet code extends the basic set of GnomeCanvasItems
available by implementing a number of specialized items.   

\begin{itemize}

	\item {Bar item:} This one is used to draw the columns and
	rows indicators.  

	\item {Cursor item:} Various number of cursors are supported
	in Gnumeric: input cursors, editing cursors, marked regions,
	walking-ant cursors.  They can have different sizes and than
	one cursor can be active at any point.

	Making the cursor a CanvasItem simplifies and centralizes the
	event handling for the cursors.

	\item {Grid item:} This is the actual spreadsheet display
	code.  It is the most complex CanvasItem in Gnumeric as it
	deals with the spreadsheet repaint code and the cell repaint
	code.

	\item {Comment item:} Cells in Gnumeric can have comments
	attached to them.   

\end{itemize}

All of the above components are tied together inside a SheetView GTK+
Widget.  A side effect of using the GNOME Canvas as the foundation for
the spreadsheet display engine is that simple drawing features can be
added to Gnumeric and embedding of arbitrary widgets inside a
spreadsheet is simple.

The support for embedding widgets is specially interesting, as the
GNOME document model [baboon] uses a special widget to embed documents
that were created by other programs. 

At the point of this writing, the basic drawing engine in Gnumeric is
being replaced with Havoc Pennington's GNOME EzPaint [ezpaint]
library: The EzPaint library implements a sample drawing program on
top of any GnomeCanvas derived object by plugging it into the
framework.  

\subsection{Display code organization.}

The top-level object managed by Gnumeric is the Workbook.  A workbook
contains one or more Sheet objects.  Sheet objects in turn are the
information repository in Gnumeric:  they contain all of the
information required to render the Spreadsheet on the screen but they
do not actually deal with any display issues. 

Each Sheet object in turn contains a SheetView object which is a
visual representation of the Sheet contents in a way familiar to the
user: it contains three canvas objects: one for the column headings,
one for the row headings and finally a GnumericSheet object which is
the container for the actual data display.

The GnumericSheet object in turn contains one ItemGrid canvas object
which is bound to the container Sheet.  A variable number of cursors,
drawing objects and cell comments are contained inside this object.

The controller-view model for the Gnumeric internals was adopted to
allow Gnumeric to become a Bonobo component.  The Bonobo architecture
would require from its components to provide various display views for
its information on demand.

\section{Numbers and formats.}

Gnumeric will render the numbers based on a format string.  This
format string is the same format string supported by Microsoft's Excel
product. 

The formatting codes allow the user to display a number in many ways:
The format code can include verbatim strings that will be inserted in
the output (for example, the currency denomination);  it also allows
to control precisely the precision and style for the number
displayed. 

An interesting feature of the formatting codes is that they support a
mode in which the number will be rendered as a date or a time
representation: this is a powerful mechanism as it simplifies the task
of date manipulation in the spreadsheet.

Dates are represented internally as the number of days since the first
day of 1900 (it should be noted that Microsoft got this wrong two
times, so Gnumeric needs to support --for the sake of Excel
compatibility-- two broken representations for numbers plus the
correct representation and this is sadly a global for the workbook).

Times are represented as a fraction of the day.  For example: 0.0
represents the beginning of the day; 1.0 represent the end of the day
and 0.5 represents noon. 

\subsection{Input Parsing.}

To simplify the task of entering dates or any other data, Gnumeric
implements a very slick trick: when the user enters any data into a
cell, it will try to match the entered text as a number plus any of
the defined formats in Gnumeric.

This input parsing engine will convert strings like ``10-Jan'' 
into a number plus a format.  In 1999, the actual number stored for
the above example would be: 36230. 

Internally, Gnumeric will generate a regular expression that describes
all the possible matches a number formatting code would render.
For each regular expression, Gnumeric will keep a parallel description
of what each match represents: a tag that describes how to convert the
matching input to a number.

The user input is thus processed against all of these regular
expressions known to Gnumeric and if a match is found the engine will
use the parallel description to create a number from the input and bind
to it the specified format name.

\section{Resource management.}

There are a number of resources that are managed by Gnumeric.  The
goal is to minimize the memory usage and the system resources used by
Gnumeric.   Resources include: pre-compiled number format
descriptions, display colors, fonts, stipple patterns and rendering
styles. 

To achieve this, all of the Gnumeric allocated resources are
considered read-only resources and are always reference counted.  The
rule is simple: the user is forbidden from making changes to a
Gnumeric-managed resource;  when a resources is used in a new place
its reference count should be incremented;  when the resource is no
longer used its reference count should be decremented. 

When a modified version of a resource is required, a copy of it must
be obtained by the resource manager.  The resource manager, in turn
will try to find an matching version for the requested in the list of
allocated resources, and if a match is found, a pointer to it will be
returned and the reference count will be incremented.


\section{Information storage.}

Information in Gnumeric is stored in Cell structures.  These Cell
structures are either attached to a Sheet object or are in the
clipboard.  No Cell is ever allowed to live anywhere else.  Period.
Cells living in the clipboard are particularly boring Cells and they
are usually just in-transit Cells, so we won't bother about them.

The other kind of Cells are those that live on the domain of a Sheet.
The Sheet structure keeps a list of all of the Cells in a hashed table
(hashed by column and row number in case you were wondering).  

The Sheet also keeps track of the column and row information (the
dimensions of it, and only if the column and row actually exists).
Gnumeric conveniently keeps the row information as doubly linked list
which hangs from the column information.  Each row contains in turn
all of the Cells for that row in doubly linked list form.

This is by no mean inneficient, because those lists are barely walked
from beginning to end.  Also, Cells in turn keep pointers to their
row and column information structures.  So it all happens to be very
efficient. 

Every cell containing a valid expression is kept on a list of
cells with expressions. 

As with other resources in Gnumeric, all of the strings contained on
the Sheet objects are reference counted.  A copy of all of the strings
used in the Sheet objects is maintained in an adaptive global hash
table.  The per-cell cost is about 60 bytes in the best case scenario
and differs slightly as the cell differentiates from existing cells.

Expressions are parsed into a tree structure and are stored in this
format.  This means that the originally entered text is lost at this
stage, but from initial experiments there is no reason to keep this
entered text format at all.  Each note of the tree structure is
reference counted.  Currently this is not used to minimize memory
usage although it is the plan, but rather it is used to allow quick
cell formula duplication.



\section{The computation engine.}

\subsection{Number evaluation.}

When an expression is entered into a Sheet, Gnumeric parses the
expression and if the expression is valid a tree representation of the
expression is attached to that Cell.  Currently Gnumeric uses the
integer and double data types from the C language to perform all of
its computations.  It is planned to replace this engine with the
Guile[guile] value system, which will provide a more powerful
spreadsheet. 


Recursive references are sometimes useful in a spreadsheet to compute
functions that converge to a number.  The computation engine can be
configured to iterate a given number of times over recursive cell
definitions or until the numbers converge within a constraint.  To
achieve this, Gnumeric keep a generation tag both on the Workbook and
on each Cell on the spreadsheet.  This tag is used on the Cell to keep
control on the re-computation.

\subsection{Re-computation.}

The re-computation in Gnumeric works in the following way: when an
expression has been successfully parsed the tree is examined for any
references to cells.  For each cell reference or cell range reference,
the code keeps track of this region of ``interest'' (internally called
a DependencyRange).  

When the contents of a cell cell in the spreadsheet is modified, the
location of this cell is looked up in the regions of interest hash
table.  If a match is found, then the cell is put on the queue of
pending re-computations. 

To actually perform the re-computation, the following steps take place:

\begin{itemize}

	\item the workbook generation counter is incremented.
	
	\item each cell on the re-computation queue is pulled from the
	queue and evaluated.

	\item A list of every cell depending on the computed cell is
	retrieved and if the dependency cell generation counter does
	not match the current workbook generation, the cell is added
	to the queue of pending re-computations.

\end{itemize}

Forcing a re-computation on the workbook is just a matter of queuing the
list of known formulas and invoking the pending computation routine. 

Various ways exist to improve the re-computation mechanism by adding
various sorts of lazy re-computation based on various criteria.
Unfortunately some of the most interesting methods that could be
implemented are patented.


\section{Plug-in setup.}

At start up Gnumeric loads any shared libraries that are installed in
the Gnumeric library directory.  These shared libraries have full
access to the Gnumeric internals and can hook themselves into various
spots inside the code.  

Plug-ins exist for various scripting languages (described in the next
section) as well as extended collections of functions for domain
specific applications. 

Plug-ins that depend on being linked into the address space of
Gnumeric to work are required to be licensed under a license
compatible with the GPL.  


\section{Scripting.}

There are a number of plug-ins written for Gnumeric that allow the user
to extend Gnumeric with a scripting language.  At the time of this
writing, plug-ins for Guile, Perl and Python exist.  The Perl and
Python plug-ins are a bit limited as they only allow new functions to
be defined in Perl and Python.  The Guile plug-in is better integrated
with the Gnumeric core and allows more powerful scripts to be created
with it. 

Guile provides various interesting features like big integers, precise
and imprecise arithmetic and complex numbers.  Besides these
advantages, the existence of various Scheme-based symbolic
computation, physics packages and advanced mathematical packages make
the Guile integration an exciting extension. 

\section{Automation.}

Gnumeric uses the ORBit[orbit] CORBA implementation and the GNOME
setup for object activation to export its functionality and allow it
to be used as a reusable component in other applications.

Gnumeric exposes a number of interfaces to the world.  A factory
interface to create new instances of Gnumeric and various interfaces
to control Gnumeric in a number of ways.  Any application on the GNOME
desktop can launch Gnumeric and control Gnumeric completely as if the
user were actually performing the actions by means of a high level
interface to it.  Scripting languages with CORBA support and any other
CORBA-enabled application on the desktop can make use of Gnumeric in
this way. 

Gnumeric exposes a number of interfaces to the world.  Besides the
Bonobo interfaces, Gnumeric can be seen as:

\begin{itemize}

	\item {\tt GNOME::Table}  Applications will consider any of
	Gnumeric's data to be a Table.  Just like a database, and they
	do not need to care about any spreadsheet specific features.

	\item {\tt GNOME::Spreadsheet}  This is a generic spreadsheet
	interface and provides access to functions that are supposed
	to be available on most spreadsheets (insert and remove
	columns and rows; set cell properties;  define plotting
	information and items like this).

	\item {\tt GNOME::Gnumeric}  This is a more specialized
	version of the {\tt GNOME::spreadsheet} interface.  This
	provides access to features that are unique to Gnumeric.

\end{itemize}

\section{The development.}

Advances in software development techniques, software engineering and
faster machines being available to developers were enabling
technologies to create Gnumeric.  The GNU Image Manipulation Program
(GIMP[gimp]) and the GTK+[gtk] toolkit projects were specially
motivating: the core and first releases of the GIMP program was
written by two programmers in a relatively short time.

The programming style, architecture and special interest in program
correctness on the GIMP and GTK+ have motivated the code base for
Gnumeric. 

The second public release of Gnumeric took place three months after
the project started and consisted of approximately 25,000 lines of
code.  By the time it was releases it was already better in terms of
the computation engine, display engine, the plug-in architecture to
any other free software offering in this area.

The approach used in Gnumeric showed a number of important things:

\begin{itemize}
	
	\item {Code correctness pays:} Using pre-conditions widely
	helped find bugs earlier in the coding.

	\item {Rapid development:} The GTK+ toolkit and GNOME
	libraries provide a powerful framework for building commercial
	grade applications with little effort. 

	\item {Big projects do not need lots of contributors:}
	Gnumeric was written mostly by a single person during the
	course of two months.  Contributors to free software can
	usually help with many of the details in providing a complete
	application, so it is not worth worrying about those at the
	beginning.  

	For example, various plug-in modules were written by
	contributors after the initial code release, including
	language bindings to various scripting languages.

	The core of the Excel 97 importing code was written in two
	weeks by a two-person team. 

	\item {Clean interfaces and inline documentation:} The
	Gnumeric code exposes its inner API in a very clean fashion
	and it uses inline documentation to describe the API.

	This allows plug-ins to be written with little effort, and the
	plug-in author does not need to understand the whole code base
	before contributing code.

	A subset of the internal API is exposed trough CORBA to the
	GNOME desktop.

	\item {C is by no means dead for Graphical User Interface
	development:} All of the Gnumeric core has been written in C.
	Many people believe that the C language is not adequate for
	GUI programming and that no complex application can be written
	with it.

	GNOME and GTK+ provide an C-based object oriented framework
	for GUI application development.  By using C we use the
	low-level common denominator in Unix languages and we can
	easily provide bindings for almost every language available on
	UNIX systems \cite{gnome2}

	It is the author belief that a share of the problems people
	have with C are not really with the language but rather with
	the way C is taught in books and by C language instructors.  

	A focus on correctness and clean coding practices would help
	the language enormously.  Learning about Emacs Emacs Dynamic
	completion or vi dynamic completion also helps (due to our
	long identifier names).
	
\end{itemize}

\section{Other Features.}

Gnumeric implements various other features found on proprietary
spreadsheets.  One of the most interesting pieces of code is the
auto-filling code.

\subsection{Automatic cell fill.}

The principle behind automatic cell filling is very simple: The user
should be able to fill a region of cells by pulling the cell selector
handle and dragging it to fill a selected region.

Automatic filling will take into consideration the currently selected
region of cells to compute the values for the automatic filling: if a
single cell is selected, then the automatic-fill delta is set to one.
If the selected cell range is bigger than that, then the delta is set
to the difference between the last value and the previous value.

Computing the values for numbers is fairly simple: the number is
already stored in the cell.  Since dates and hours are stored
internally as some number plus a format, this automatically takes care
of automatic filling for dates and hours. 

For non-numerical values, Gnumeric will try to match the value against
a set of string tables: This allows automatic filling to work with a
list of words (weekdays and month names for example).

Gnumeric attempts finally to match a number at the beginning or at the
end of the cell text.  If it finds a match, it uses that number for
the automatic filling and during the automatic filling it will replace
the number with the value computed.  This handles cases for automatic
filling string like ``Product 1''. 

If all of the above methods fail, Gnumeric will just copy the cell
values. 

\section{Future development.}

    The Gnumeric team considers the lack of important spreadsheet
features in the code a bug, and thus they will get fixed eventually.
Besides the regular features that are expected to exist in a
spreadsheet, there are various areas in which Gnumeric will be moved
in the following months: 

\begin{itemize}

	\item {Expression optimization at compile time:} Currently
	expressions are stored in the same way that the user entered
	them.  Simple optimizations can be performed to the expression
	once the expression tree has been built:  common subexpression
	elimination and constant folding should be easily
	implemented.   

	\item {Finish Bonobo integration:} Currently Gnumeric does not
	support any of the Bonobo interfaces for document embedding
	and in-place activation.  Bonobo, Gnumeric and the rest of the
	productivity applications in GNOME are evolving hand in hand
	to provide an integrated set of tools for GNOME. 

	\item {Guppi Graphics integration:} Guppi is the plotting
	system of GNOME.  The plotting system is commonly viewed as
	tightly bound to a spreadsheet.  In GNOME the plotting system
	and the spreadsheet are separate and the interactions between
	them is achieves by two sets of interfaces:  The Bonobo
	interfaces for document embedding and in-place editing and the
	Table and Plot interfaces (the former provided by Gnumeric and
	the later provided by Guppi).  

	This is another step towards component reuse and integration. 

	\item {Constraint solver:} Besides the obvious uses of a
	constraint solver, a constraint solver is required for a nicer
	Guppi integration. 

	Guppi has been modified to allow user intervention and data
	modification with the mouse.  Changes on the plot to a
	computed formula will trigger the constraint solver to find
	appropriate values for the domain of a function. 

	\item {Using Guile math engine for computation:} Gnumeric
	currently uses doubles and integers from C to perform its
	computations.  We want to go beyond this and provide access to
	the powerful Guile value system to the Gnumeric user.  

	Integration of this engine

	\item {Printing:} Finishing the Printing support for Gnumeric.
	Currently, there is an experimental branch of Gnumeric that
	includes printing support using the gnome-print printing
	architecture.

\end{itemize}

\section{File I/O.}

Gnumeric has a framework to plug file loading and file saving
routines.  File formats are usually implemented as plug-in modules to
Gnumeric. 

At the time of this writing, Gnumeric currently supports four file
formats: an XML-based file format, the Excel97 file format, a text
importer and a comma separated value format.

The Gnumeric native file format is based on the XML standard.  Daniel
Veillard at the W3 consortium designed the Gnumeric native file
format.  Since the XML file format tends to be a bit verbose, Gnumeric
by default compresses the files with the libz[libz] library to
minimize space consumption.  This format supports all of the features
supported in Gnumeric. 

The Excel support was written by Michael Meeks and Bruno Unna from
Microsoft published documentation for both the BIFF format used by
Excel and the OLE structured storage file format.  A side effect of
the Excel97 work has been that GNOME now has a library for dealing
with Structured Storage OLE2 files.

\section{Acknowledgements.}

The hackers of the GNOME project which have provided valuable input
during the design of this code.  Specially the Gnumeric hackers which
have provided many features to the code base.

\section{Availability.}

Up to date information about Gnumeric, its availability and latest
features is available from the Gnumeric web page at: 

\begin{center}\small
{\texttt {http://www.gnome.org/gnumeric}}
\end{center}

The Gnumeric spreadsheet is available from the GNOME FTP server:

\begin{center}\small
	{\texttt{ftp://ftp.gnome.org}}
\end{center}

and its mirrors.  The development versions of Gnumeric are available
from the GNOME CVS server and its anonymous-login replicas as the
``gnumeric'' module.  

A number of experimental features of Gnumeric are usually kept on
separate branches on the GNOME CVS repository.  The GNOME team
provides Bonsai and LXR access to the GNOME source code in the web
site at:

\begin{center}\small
{\texttt {http://cvs.gnome.org/}}
\end{center}

\bibliography{gnumeric}

\noindent
[sfaq] The Spreadsheet FAQ.

\noindent
[baboon] The GNOME document model.

\noindent
[ezpaint] The GNOME EzPaint library, Havoc Pennington.

\noindent
[GIMP] The GNU Image Manipulation Program, http://www.gimp.org

\noindent
[GTK+] The GIMP Toolkit, http://www.gtk.org

\noindent
[gnufree] http://www.gnu.org/philosophy/free-sw.html

\noindent
[guile] http://www.gnu.org

\noindent
[libz] http://www.cdrom.com/pub/infozip/zlib

\noindent
[omg] The Object Management Group, http://www.omg.org

\noindent
[orbit] The ORBit CORBA 2.2 implementation,
http://www.labs.redhat.com/orbit 

\end{document}
